---
title: "Querying GeneNetwork2 for Model Organism QTL Data"
author: "Brian Gural, Todd Kimball, Anh Luu, Christoph D. Rau"
date: "`r Sys.Date()`"
output:
    rmarkdown::html_vignette:
        toc: true
        toc_depth: 3
vignette: >
  %\VignetteIndexEntry{Querying GeneNetwork2 for Model Organism QTL Data}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  eval = FALSE
)
```

# Overview

This vignette demonstrates how to query the [GeneNetwork2](https://genenetwork.org)
REST API to retrieve eQTL and phenotype data from model organism genetic reference
populations (GRPs), and how to integrate those results into a `locusPackRat` project
as additional evidence layers.

GeneNetwork2 hosts QTL mapping results, expression data, and phenotype records for
several GRPs, including:

- **BxD recombinant inbred lines** (mouse): extensive eQTL, pQTL, and phenotype data
- **Diversity Outbred (DO)** (mouse): high-resolution QTL mapping panels
- **Hybrid Mouse Diversity Panel (HMDP)**: cardiovascular and metabolic trait data
- **Other organisms**: rat (HxB/BxH), Drosophila, and Arabidopsis panels

Unlike Open Targets, which focuses on human GWAS data, GeneNetwork2 provides
population-specific molecular QTL data that is directly relevant to model organism
studies. This makes it a valuable complement to the human-centric annotations
retrieved by `queryOpenTargets()` and `queryOpenTargetsQTL()`.

All code chunks in this vignette are set to `eval=FALSE` to avoid API dependency
during package builds. Users should run the code interactively.

# Prerequisites

```{r libraries}
library(locusPackRat)
library(data.table)
library(httr)
library(jsonlite)
```

# GeneNetwork2 API Basics

The GeneNetwork2 REST API is available at `https://genenetwork.org/api/v_pre1/`.
Key endpoints include:

- **`/species`**: List available species
- **`/groups/{species}`**: List available populations (e.g., BXD, DO)
- **`/datasets/{group}`**: List available datasets for a population
- **`/traits/{dataset}`**: List traits within a dataset
- **`/trait/{dataset}/{trait_id}`**: Get detailed trait information

```{r api_base}
# Base URL for GeneNetwork2 API
gn2_base <- "https://genenetwork.org/api/v_pre1"
```

# Example 1: BxD eQTL Data

## List Available BxD Expression Datasets

The BxD panel has dozens of expression datasets spanning multiple tissues
(brain, liver, heart, kidney, etc.) and platforms (microarray, RNA-seq).

```{r list_bxd_datasets}
# Fetch all datasets for the BXD group
bxd_datasets <- fromJSON(content(
  GET(paste0(gn2_base, "/datasets/BXD")),
  as = "text", encoding = "UTF-8"
))

# Filter for expression datasets (mRNA / RNA-seq)
expression_datasets <- bxd_datasets[grepl("mRNA|RNA-seq|expression",
                                          bxd_datasets$tissue, ignore.case = TRUE), ]

# Preview available expression datasets
head(expression_datasets[, c("id", "name", "tissue", "description")])
```

## Query Traits for a Specific Gene

To check whether a candidate gene has eQTL evidence in the BxD panel,
query traits by gene symbol within a specific dataset.

```{r query_gene_traits}
# Example: query Cisd2 expression traits in BXD heart dataset
# Dataset ID will vary — use the listing above to find the correct one
dataset_name <- "HC_M2_0606_P"  # Example: heart dataset

# Search for traits matching a gene symbol
trait_search_url <- paste0(gn2_base, "/traits/", dataset_name)
heart_traits <- fromJSON(content(
  GET(trait_search_url),
  as = "text", encoding = "UTF-8"
))

# Filter for a specific gene of interest
gene_of_interest <- "Cisd2"
cisd2_traits <- heart_traits[grepl(gene_of_interest,
                                    heart_traits$symbol, ignore.case = TRUE), ]

if (nrow(cisd2_traits) > 0) {
  cat("Found", nrow(cisd2_traits), "probes/traits for", gene_of_interest, "\n")
  print(cisd2_traits[, c("id", "symbol", "description")])
}
```

## Retrieve Trait Values Across BxD Strains

Once you have identified a trait of interest, retrieve the expression values
across all BxD strains.

```{r get_trait_values}
# Get expression values for a specific trait across BxD strains
trait_id <- cisd2_traits$id[1]  # Use the first matching probe

trait_url <- paste0(gn2_base, "/trait/", dataset_name, "/", trait_id)
trait_data <- fromJSON(content(
  GET(trait_url),
  as = "text", encoding = "UTF-8"
))

# Convert to data.table for locusPackRat integration
if (!is.null(trait_data$sample_data)) {
  strain_expression <- data.table(
    strain = names(trait_data$sample_data),
    expression = unlist(trait_data$sample_data)
  )
  cat("Expression data for", nrow(strain_expression), "BxD strains\n")
  head(strain_expression)
}
```

## Run QTL Mapping via GEMMA

GeneNetwork2 can perform on-the-fly QTL mapping using GEMMA (Genome-wide
Efficient Mixed Model Association). This returns LOD scores across the genome.

```{r run_gemma}
# Request GEMMA mapping for a trait
# Note: this endpoint may take some time to return results
gemma_url <- paste0(gn2_base, "/mapping?trait_id=", trait_id,
                    "&db=", dataset_name,
                    "&method=gemma")

gemma_results <- fromJSON(content(
  GET(gemma_url, timeout(120)),
  as = "text", encoding = "UTF-8"
))

# Convert to data.table
if (length(gemma_results) > 0) {
  eqtl_scan <- data.table(
    chr = gemma_results$chr,
    pos = gemma_results$Mb * 1e6,  # Convert Mb to bp
    lod = gemma_results$lod_score,
    gene_symbol = gene_of_interest,
    source = "GeneNetwork2_BxD"
  )

  # Filter for significant peaks (e.g., LOD > 3)
  significant_eqtl <- eqtl_scan[lod > 3]
  cat("Found", nrow(significant_eqtl), "significant eQTL positions\n")
}
```

# Example 2: DO Phenotype Data

The Diversity Outbred (DO) population provides higher mapping resolution
than the BxD panel due to its greater genetic diversity.

## List Available DO Datasets

```{r list_do_datasets}
# Fetch datasets for the DO group
do_datasets <- fromJSON(content(
  GET(paste0(gn2_base, "/datasets/DO")),
  as = "text", encoding = "UTF-8"
))

# Preview
head(do_datasets[, c("id", "name", "tissue", "description")])
```

## Query DO Phenotype Traits

```{r query_do_phenotypes}
# Example: query a DO phenotype dataset
do_dataset_name <- do_datasets$name[1]  # Use first available

do_traits <- fromJSON(content(
  GET(paste0(gn2_base, "/traits/", do_dataset_name)),
  as = "text", encoding = "UTF-8"
))

# Preview available traits
if (nrow(do_traits) > 0) {
  head(do_traits[, c("id", "symbol", "description")])
}
```

# Integrating GeneNetwork2 Results into locusPackRat

## Format eQTL Results for addRatTable()

GeneNetwork2 QTL results can be added to a locusPackRat project using
`addRatTable()` with either gene-level or point-level linking.

### Gene-Level Integration

Use `link_type = "gene"` when you have per-gene summary statistics
(e.g., peak LOD score, cis/trans classification).

```{r integrate_gene_level}
# Summarize eQTL results per gene (e.g., maximum cis-eQTL LOD)
# In practice, you would loop over multiple candidate genes
eqtl_summary <- data.table(
  gene_symbol = c("Cisd2", "Pdlim5", "Manba"),
  bxd_max_lod = c(5.2, 3.1, 4.8),
  bxd_peak_chr = c("3", "3", "3"),
  bxd_peak_pos = c(135000000, 142000000, 138000000),
  bxd_cis_eqtl = c(TRUE, FALSE, TRUE),
  gn2_dataset = rep("HC_M2_0606_P", 3)
)

# Add to locusPackRat project
addRatTable(
  data = eqtl_summary,
  table_name = "bxd_eqtl",
  abbreviation = "bxd",
  link_type = "gene",
  link_by = "gene_symbol",
  project_dir = "my_qtl_project"
)
```

### Point-Level Integration

Use `link_type = "point"` when you have position-level QTL scan data
that should be mapped to overlapping genes/regions.

```{r integrate_point_level}
# Use the full scan data from GEMMA
# Ensure columns match the expected format: chr, pos (or start/end)
scan_for_project <- eqtl_scan[, .(
  chr = chr,
  start = as.integer(pos),
  end = as.integer(pos),
  lod = lod,
  gene_symbol = gene_symbol,
  source = source
)]

addRatTable(
  data = scan_for_project,
  table_name = "bxd_eqtl_scan",
  abbreviation = "bxds",
  link_type = "region",
  link_by = "chr,start,end",
  project_dir = "my_qtl_project"
)
```

# Alternative: GNapi R Package

The [GNapi](https://github.com/kbroman/GNapi) R package by Karl Broman
provides a more structured R client for the GeneNetwork2 API. If you
prefer a packaged interface:

```{r gnapi_example}
# Install GNapi
# remotes::install_github("kbroman/GNapi")

library(GNapi)

# List available species
species <- list_species()

# List BxD datasets
bxd_datasets <- list_datasets("BXD")

# Get trait data
trait_data <- get_trait("HC_M2_0606_P", "your_trait_id")
```

# Tips and Considerations

1. **API Rate Limits**: GeneNetwork2 may impose rate limits. Add `Sys.sleep(1)`
   between requests when querying many genes.

2. **Genome Build**: GeneNetwork2 positions are typically in GRCm38/mm10 for
   mouse data. If your locusPackRat project uses mm39, convert coordinates
   using `rtracklayer::liftOver()` before integration (see the liftOver
   section in the main workflow vignette).

3. **Dataset Selection**: Choose tissue-appropriate datasets. For a cardiac
   study, prioritize heart expression datasets over brain or liver.

4. **Cis vs Trans**: A cis-eQTL (peak near the gene itself) is generally
   stronger evidence for candidacy than a trans-eQTL.

5. **Cross-Validation**: Compare GeneNetwork2 eQTL results with your own
   study-specific eQTL data and Open Targets QTL data for convergent evidence.

# Session Info

```{r session_info}
sessionInfo()
```
