---
title: "locusPackRat: Project-Based Genomic Analysis Workflow"
author: "Brian Gural, Todd Kimball, Anh Luu, Christoph D. Rau"
date: "`r Sys.Date()`"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{locusPackRat Workflow}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.width = 7,
  fig.height = 5,
  warning = FALSE,
  message = FALSE
)
```

## Introduction

The `locusPackRat` package provides a project-based workflow for genomic analyses with persistent data storage and integrated visualization. Instead of repeatedly passing data between functions, you create persistent projects that store genes or genomic regions along with any supplementary data.

This vignette demonstrates the complete workflow using real genomic data from chromosome 5.

## Installation

```{r install, eval=FALSE}
# Install from GitHub
devtools::install_github("RauLabUNC/locusPackRat")
```

## Load Required Libraries

```{r libraries}
library(locusPackRat)
library(data.table)
library(jsonlite)
```

## Complete Workflow Example

This example demonstrates the full workflow using real genomic coordinates from chromosome 5.

### Step 1: Load and Prepare Genomic Data

First, we'll load real genomic coordinates and subset to chromosome 5 for this demonstration:

```{r load_data}
# Load test data (included with package)
mouse_coords <- fread(system.file("extdata", "mouse_coords_mm39.csv",
                                  package = "locusPackRat"))
human_coords <- fread(system.file("extdata", "human_coords_hg38.csv",
                                  package = "locusPackRat"))

# Subset to chromosome 5
chr5_coords <- mouse_coords[chr == "5"]
cat("Found", nrow(chr5_coords), "genes on chromosome 5\n")

# Create primary gene data from real coordinates
primary_genes <- data.table(
  gene_symbol = chr5_coords$gene_symbol,
  ensembl_id = chr5_coords$ensembl_id,
  chr = chr5_coords$chr,
  start = chr5_coords$start,
  end = chr5_coords$end
)

# Save gene symbols before they get modified by initPackRat
gene_symbols <- chr5_coords$gene_symbol
n_genes <- nrow(primary_genes)
```

### Step 2: Initialize a locusPackRat Project

Create a new project directory structure with your genomic data:

```{r init_project}
# Initialize project with chromosome 5 genes
initPackRat(
  data = primary_genes,
  mode = "gene",
  species = "mouse",
  genome = "mm39",
  project_dir = "chr5_analysis",
  force = TRUE
)
```

This creates a `.locusPackRat` directory with:
- `input/genes.csv` - Your input genes with coordinates
- `input/orthology.csv` - Cross-species orthology information
- `config.json` - Project configuration
- `supplementary/` - Directory for additional data tables

### Step 3: Add Supplementary Data Tables

Now we can add various types of supplementary data to our project:

#### Expression Data

```{r add_expression}
# Create expression data (using simulated values for demonstration)
set.seed(12)
expression_data <- data.table(
  gene_symbol = gene_symbols,
  brain = round(abs(rnorm(n_genes, 100, 50)), 1),
  liver = round(abs(rnorm(n_genes, 80, 40)), 1),
  heart = round(abs(rnorm(n_genes, 90, 45)), 1),
  kidney = round(abs(rnorm(n_genes, 70, 35)), 1)
)
expression_data[, high_expression := pmax(brain, liver, heart, kidney, na.rm = TRUE) > 120]

# Add to project
addRatTable(
  data = expression_data,
  table_name = "expression",
  link_type = "gene",
  link_by = "gene_symbol",
  project_dir = "chr5_analysis"
)
```

#### Phenotype Data

```{r add_phenotypes}
# Create phenotype associations
phenotype_data <- data.table(
  gene_symbol = sample(gene_symbols, min(60, n_genes)),
  phenotype = sample(c("lethal", "abnormal behavior", "metabolic", "immune", "growth"),
                    min(60, n_genes), replace = TRUE),
  p_value = 10^(-runif(min(60, n_genes), 0.5, 8)),
  effect_size = round(rnorm(min(60, n_genes), 0, 1), 2)
)
phenotype_data[, significant := p_value < 0.05]

addRatTable(
  data = phenotype_data,
  table_name = "phenotypes",
  link_type = "gene",
  link_by = "gene_symbol",
  project_dir = "chr5_analysis"
)
```

#### QTL Statistics

```{r add_qtl}
# Add QTL mapping results
qtl_stats <- data.table(
  gene_symbol = gene_symbols,
  lod_score = round(runif(n_genes, 0, 12), 2),
  p_value = 10^(-runif(n_genes, 0.5, 8)),
  fold_change = round(rnorm(n_genes, mean = 0, sd = 2), 2)
)
qtl_stats[, significant := p_value < 0.05]

addRatTable(
  data = qtl_stats,
  table_name = "qtl_statistics",
  link_type = "gene",
  link_by = "gene_symbol",
  project_dir = "chr5_analysis"
)
```

#### Human Orthology

```{r add_orthology}
# Load orthology mapping
orthology_map <- fread(system.file("extdata", "orthology_mapping.csv",
                                   package = "locusPackRat"))

# Get matching orthologs for our chr5 genes
ortho_matches <- orthology_map[mouse_gene_symbol %in% gene_symbols]

if (nrow(ortho_matches) > 0) {
  # Join with human coordinates
  ortho_data <- merge(ortho_matches, human_coords,
                      by.x = "human_gene_symbol", by.y = "gene_symbol",
                      all.x = TRUE)
  # Prepare orthology table
  ortho_data <- ortho_data[, .(
    gene_symbol = mouse_gene_symbol,
    human_gene_symbol = human_gene_symbol,
    human_ensembl_id = ensembl_id,
    human_chr = chr,
    human_start = start,
    human_end = end
  )]

  addRatTable(
    data = ortho_data,
    table_name = "human_orthology",
    link_type = "gene",
    link_by = "gene_symbol",
    project_dir = "chr5_analysis"
  )
}
```

### Step 4: List Available Tables

You can check what supplementary tables are available in your project:

```{r list_tables}
tables <- listPackRatTables("chr5_analysis")
cat("Available supplementary tables:\n")
for (table in tables) {
  cat("  •", table, "\n")
}
```

### Step 5: Generate Output Files

Now we can generate various output files that automatically integrate all our data:

#### Simple CSV Output

```{r generate_csv, eval=FALSE}
generateGeneSheet(
  format = "csv",
  output_file = "chr5_complete.csv",
  include_supplementary = TRUE,
  project_dir = "chr5_analysis"
)
```

#### Filtered Output

```{r generate_filtered, eval=FALSE}
# Only significant QTL hits
generateGeneSheet(
  format = "csv",
  output_file = "chr5_significant.csv",
  filter_expr = "p_value < 0.05 & lod_score > 5",
  include_supplementary = TRUE,
  project_dir = "chr5_analysis"
)
```

#### Multi-Tab Excel with Conditional Formatting

```{r generate_excel}
# Pick some genes to highlight
genes_to_highlight <- head(gene_symbols[!is.na(gene_symbols)], 5)
cat("Highlighting:", paste(genes_to_highlight, collapse = ", "), "\n")

generateGeneSheet(
  format = "excel",
  output_file = "chr5_analysis_multisheet.xlsx",
  split_by = "criteria",
  split_criteria = list(
    "All_Chr5" = "TRUE",
    "High_LOD" = "lod_score > 7",
    "Significant_QTL" = "!is.na(lod_score) & p_value < 0.05",
    "High_Expression" = "high_expression == TRUE",
    "With_Phenotype" = "!is.na(phenotype)",
    "Human_Ortholog" = "!is.na(human_gene_symbol)"
  ),
  highlight_genes = genes_to_highlight,
  include_supplementary = TRUE,
  project_dir = "chr5_analysis"
)
```

This creates an Excel workbook with:
- Multiple tabs based on filtering criteria
- Yellow highlighting for specified genes
- Auto-sized columns
- Frozen header rows
- Professional formatting

#### Selective Table Inclusion

```{r generate_selective, eval=FALSE}
# Include only specific supplementary tables
generateGeneSheet(
  format = "excel",
  output_file = "chr5_qtl_expression.xlsx",
  include_supplementary = c("qtl_statistics", "expression"),
  project_dir = "chr5_analysis"
)
```

## Working with Region-Based Projects

The package also supports genomic region-based projects (e.g., QTL intervals):

```{r region_project, eval=FALSE}
# Define QTL regions
qtl_regions <- data.table(
  chr = c(5, 10, 12),
  start = c(10000000, 50000000, 75000000),
  end = c(15000000, 55000000, 80000000),
  trait = c("body_weight", "glucose", "cholesterol"),
  lod = c(8.5, 6.2, 7.8)
)

# Initialize region-based project
initPackRat(
  data = qtl_regions,
  mode = "region",
  species = "mouse",
  genome = "mm39",
  project_dir = "qtl_regions",
  force = TRUE
)

# Add scan data
scan_data <- data.table(
  chr = rep(5, 100),
  pos = seq(10000000, 15000000, length.out = 100),
  lod = rnorm(100, mean = 5, sd = 2) +
        ifelse(seq(10000000, 15000000, length.out = 100) > 12000000 &
               seq(10000000, 15000000, length.out = 100) < 13000000, 3, 0)
)

addRatTable(
  data = scan_data,
  table_name = "qtl_scans",
  link_type = "region",
  project_dir = "qtl_regions"
)
```

## Generating LocusZoom Plots

For region-based projects, you can generate LocusZoom-style visualizations:

```{r locuszoom, eval=FALSE}
generateLocusZoomPlot_v2(
  region_id = 1,  # First region
  scan_table = "qtl_scans",
  highlight_genes = c("Gene1", "Gene2"),
  output_file = "chr5_locuszoom.pdf",
  project_dir = "qtl_regions"
)
```

## Key Benefits of the locusPackRat Workflow

1. **Persistent Storage**: Load data once, use many times
2. **Flexible Integration**: Add any type of supplementary data
3. **Reproducible Outputs**: Generate different outputs with different parameters
4. **Multi-Species Support**: Built-in orthology mapping
5. **Excel Integration**: Professional multi-sheet workbooks with conditional formatting
6. **Project Organization**: All related data in one place

## Session Information

```{r session}
sessionInfo()
```