---
title: "locusPackRat: Project-Based Genomic Analysis"
author: "Brian Gural, Todd Kimball, Anh Luu, Christoph D. Rau"
date: "`r Sys.Date()`"
output:
    rmarkdown::html_vignette:
        toc: true
        toc_depth: 3
vignette: >
  %\VignetteIndexEntry{locusPackRat: Project-Based Genomic Analysis}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.width = 7,
  fig.height = 5,
  warning = FALSE,
  message = FALSE
)
```

# Overview

`locusPackRat` is a coordinate-based genomic analysis package for R that provides a project-based workflow with persistent data storage. The package enables users to integrate multiple data types and generate professional multi-sheet Excel workbooks for genomic analyses.

`locusPackRat` functions can be grouped into the following categories:

- **Project initialization functions:** Create and configure analysis projects for genes or genomic regions
- **Data integration functions:** Add supplementary data tables that link to genes or regions
- **Output generation functions:** Export integrated data as CSV or multi-sheet Excel workbooks

## Key Features

- **Persistent project storage:** Load data once, use many times
- **Flexible data integration:** Link any supplementary data by gene symbol, Ensembl ID, or coordinates
- **Genome-wide annotations:** Built-in coordinate files for ~60,000 mouse and ~30,000 human genes
- **Cross-species support:** Automatic orthology mapping between human and mouse
- **Professional outputs:** Multi-sheet Excel workbooks with conditional formatting

## Data Resources

The package includes complete genome-wide coordinate files generated from Ensembl Biomart:

- `mouse_coords_mm39.csv`: ~60,000 mouse genes (GRCm39)
- `mouse_coords_mm10.csv`: ~38,000 mouse genes (GRCm10)
- `human_coords_hg38.csv`: ~30,000 human genes (GRCh38)
- `human_coords_hg19.csv`: ~24,000 human genes (GRCh19)

## Data Sources Used in This Vignette

Throughout this vignette, we draw on the following data types:

- **Gene coordinates and orthology**: Built-in Ensembl Biomart coordinate files and human-mouse orthology tables
- **MouseMine phenotype annotations**: Gene-phenotype associations from the Mouse Genome Informatics (MGI) database, queried via `queryMouseMine()`
- **Open Targets disease associations**: Human gene-disease links, genetic constraints (gnomAD), and drug tractability data, queried via `queryOpenTargets()`
- **Study-specific data**: User-supplied tables (e.g., differential expression, eQTL, mutations) added via `addRatTable()`

For molecular QTL data from Open Targets (eQTL, pQTL, sQTL), see the `region_qtl_opentargets` vignette. For model organism QTL data from GeneNetwork2, see the `genenetwork_qtl` vignette.

# Installation

```{r install, eval=FALSE}
# Install from GitHub
#devtools::install_github("RauLabUNC/locusPackRat")

#Or, install locally
#devtools::install_local("Path/To/File/locusPackRat.zip")

# Load required libraries
library(locusPackRat)
library(data.table)
library(jsonlite)
```

```{r libraries, echo=FALSE}
library(locusPackRat)
library(data.table)
library(jsonlite)
```

# Gene Mode Analysis

Gene mode is designed for analyzing gene lists from RNA-seq or other gene-centric studies. In this mode, you provide a data.frame with a `gene_symbol` column and any additional columns of per-gene data (e.g., log2 fold changes, p-values). `initPackRat()` maps these gene symbols to genomic coordinates and creates the project directory structure. See below for Region Mode, designed for analysis of a chromosome region.

## Initialize Project with Gene List

The first step is to call `initPackRat()` with your gene data. The `mode = "gene"` argument tells the function to expect a gene list rather than genomic intervals. The `species` and `genome` arguments determine which built-in coordinate file is used for mapping. **Important**: all data added to the project later must use the same genome build.

```{r gene_mode_init, message=TRUE}
# Prepare example RNA-seq differential expression results
set.seed(27)
rna_genes <- data.table(
  gene_symbol = c(
    c("Myc", "Tp53", "Egfr", "Vegfa", "Il6", "Tnf", "Apoe", "Lep", "Ins2", "Bdnf"),
    sample(fread(system.file("extdata", "mouse_coords_mm39.csv",
                             package = "locusPackRat"))$gene_symbol, 140)
  ),
  log2FC = round(rnorm(150, mean = 0, sd = 2), 2),
  padj = 10^(-runif(150, 0.5, 10)),
  baseMean = round(10^runif(150, 1, 5), 1),
  direction = ifelse(rnorm(150) > 0, "up", "down")
)

# Initialize project - automatically adds genomic coordinates
initPackRat(
  data = rna_genes,
  mode = "gene",
  species = "mouse",
  genome = "mm39",
  project_dir = "rnaseq_analysis",
  force = TRUE
)
```

## Add Supplementary Data

Supplementary data tables are added to the project using `addRatTable()`. Each table must contain a column that can be linked to the gene list — either `gene_symbol`, `ensembl_gene_id`, or genomic coordinates (`chr`, `start`, `end`). The `link_type` parameter specifies how the data connects to the project genes, and `link_by` identifies the linking column(s).

### Tissue Expression Atlas

Tissue expression data helps identify whether candidate genes are expressed in disease-relevant tissues. Here we simulate TPM values across six tissues and calculate tissue specificity metrics.

```{r add_tissue_expression}
# Load processed genes with coordinates
processed_genes <- fread("rnaseq_analysis/.locusPackRat/input/genes.csv")

# Add tissue expression data (TPM values)
tissue_expression <- data.table(
  gene_symbol = processed_genes$gene_symbol,
  brain = round(abs(rnorm(nrow(processed_genes), 50, 30)), 1),
  liver = round(abs(rnorm(nrow(processed_genes), 40, 25)), 1),
  heart = round(abs(rnorm(nrow(processed_genes), 45, 28)), 1),
  kidney = round(abs(rnorm(nrow(processed_genes), 35, 20)), 1),
  lung = round(abs(rnorm(nrow(processed_genes), 42, 24)), 1),
  muscle = round(abs(rnorm(nrow(processed_genes), 38, 22)), 1)
)

# Calculate tissue specificity
tissue_expression[, max_tissue := names(.SD)[max.col(.SD)],
                 .SDcols = c("brain", "liver", "heart", "kidney", "lung", "muscle")]
tissue_expression[, tissue_specific := apply(.SD, 1, function(x) max(x)/mean(x) > 2),
                 .SDcols = c("brain", "liver", "heart", "kidney", "lung", "muscle")]

addRatTable(
  data = tissue_expression,
  table_name = "tissue_expression",
  abbreviation = "te",
  link_type = "gene",
  link_by = "gene_symbol",
  project_dir = "rnaseq_analysis"
)
```

### Pathway Annotations

Pathway membership data provides functional context for candidate genes. Note that `addRatTable()` accepts any data.frame — the only requirement is that it contain a column matching `link_by` that can be joined to the project gene list.

```{r add_pathways}
# Add pathway membership data
pathways <- c("Cell cycle", "Apoptosis", "Immune response", "Metabolism",
              "Signal transduction", "DNA repair", "Protein synthesis")

pathway_data <- data.table(
  gene_symbol = sample(processed_genes$gene_symbol, 80, replace = TRUE),
  pathway = sample(pathways, 80, replace = TRUE),
  evidence = sample(c("experimental", "computational", "literature"), 80, replace = TRUE)
)

addRatTable(
  data = unique(pathway_data),
  table_name = "pathway_annotations",
 # abbreviation = "pa",
  link_type = "gene",
  link_by = "gene_symbol",
  project_dir = "rnaseq_analysis"
)
```

### Add mouseMine and Open Targets data

`queryMouseMine()` retrieves gene-phenotype associations from the Mouse Genome Informatics database, and `queryOpenTargets()` retrieves human disease associations, genetic constraint scores (gnomAD), and drug tractability data. Both functions query their respective APIs and automatically store results as supplementary tables in the project. For mouse projects, genes are mapped to human orthologs before querying Open Targets.

```{r add_mouseMine_and_Open_Targets_data_gene}
queryMouseMine(project_dir = "rnaseq_analysis")
queryOpenTargets(project_dir = "rnaseq_analysis")
```

## Examine Package Structure Before Creating Outputs

```{r gene_examine_package, message=T}
listPackRatTables(project_dir = "rnaseq_analysis", full_info = TRUE)

```

## Generate Multi-Sheet Excel Workbook

`makeGeneSheet()` merges all supplementary tables with the gene list and generates output workbooks. The `split_criteria` parameter accepts a named list of R expressions that define filtered subsets — each becomes a separate sheet in the Excel workbook. This is the key customization point: you can combine any columns from any attached table in your filter expressions.

Additional sheets can be created from any boolean (True/False) argument.
Boolean statements in R can be strung together with & (and) or | (or).
We recommend examining the package structure above with full_info = TRUE to identify column names

Note: beware of capitalization.  For whatever reason mouseMine does not capitalize any terms other than gene names, while OpenTargets is somewhat inconsistent.  If needed, more complex arguments e.g. ('Cardiac' | 'cardiac') & padj <.001 may be used.

Note 2:  If two supplemental tables share a column name, the code will append the table's abbreviation to the column name: (e.g. if the table abbreviation is 'ab' and the column is 'value', it will become 'ab_value').  It is also possible to set the prefix_mode argument below to "abbreviated", in which case all columns from tables with an abbreviation will have it appended by default.

```{r generate_gene_excel}
# Create Excel workbook with multiple filtered views
makeGeneSheet(
  format = "excel",
  output_file = "rnaseq_complete_analysis.xlsx",
  split_by = "criteria",
  prefix_mode = "collision",
    
  split_criteria = list(
    "All_DEGs" = "TRUE",
    "Upregulated" = "direction == 'up'",
    "Downregulated" = "direction == 'down'",
    "High_Confidence" = "padj < 0.001 & abs(log2FC) > 2",
    "Brain_Specific" = "max_tissue == 'brain' & tissue_specific == TRUE",
    "Immune_Genes" = "pathway == 'Immune response'",
    "mouseMine_Immune" = "grepl('immune',phenotype)"
  ),
  highlight_genes = c("Myc", "Tp53", "Egfr", "Il6", "Tnf"),
  include_supplementary = TRUE,
  project_dir = "rnaseq_analysis"
)
```

# Region Mode Analysis

Region mode is designed for analyzing genomic intervals from ATAC-seq, ChIP-seq, or QTL mapping. In this mode, you provide a data.frame with `chr`, `start`, and `end` columns defining the genomic intervals. `initPackRat()` identifies all genes overlapping these regions using the built-in coordinate files and creates the project structure.

We include two versions of Region Mode Analysis: The first for QTL mapping, the second for ATAC-seq data.

## QTL Project
### Initialize Project with Genomic Regions

```{r region_mode_init_QTL, message=TRUE}
#LocusPackRat includes an example QTL output with a single significant locus on chromsome 3.
#Here we examine two regions, one significant and the other one insignificant

QTL_peaks <- data.table(
  chr = as.character(c(3,6)),
  start=c(126967000,50000000),
  end=c(149860000,60000000),
  peak_id=c("Significant_1","Insignificant_1"),
  max_lod=c(5.64, 3.21)
)

initPackRat(
  data = QTL_peaks,
  mode = "region",
  species = "mouse",
  genome = "mm39",
  project_dir = "qtl_analysis",
  keep_pseudo=F,
  force = TRUE
)


```

### Add Supplementary Data
#### Add QTL Data and CC Founder Information

```{r add_QTL_and_Founder_Information}
QTL_file <- system.file("extdata", "sample_scan.csv", package = "locusPackRat")
QTL_Data <- fread(QTL_file)

QTL_Data <- data.table(
  chr = QTL_Data$chr,
  start = QTL_Data$pos,
  end = QTL_Data$pos,
  lod = QTL_Data$lod,
  marker = QTL_Data$marker
)

addRatTable(
  data = QTL_Data,
  table_name = "QTL_results",
  abbreviation = "qr",
  link_type = "region",
  link_by = "chr,start,end",
  project_dir = "qtl_analysis"
)

founder_file <- system.file("extdata", "sample_founders.csv", package = "locusPackRat")
founder_Data <- fread(founder_file)

addRatTable(
  data = founder_Data,
  table_name = "founder_info",
  abbreviation = "fi",
  link_type = "region",
  link_by = "chr,start,end",
  project_dir = "qtl_analysis"
)


```

#### Add Mutation Information

Non-synonymous mutation information is provided for these two loci and drawn from the Wellcome Trust Mouse Genomes Resource

```{r add NS_SNP Information}

SNP_file <- system.file("extdata", "sample_NS_SNP.csv", package = "locusPackRat")
SNP_Data <- fread(SNP_file)
SNP_Data$start <- SNP_Data$pos
SNP_Data$end <- SNP_Data$pos

addRatTable(
  data = SNP_Data,
  table_name = "NS_SNP_information",
  abbreviation = "NSi",
  link_type = "region",
  link_by = "chr,start,end",
  project_dir = "qtl_analysis"
)


```

#### Add mouseMine and Open Targets data

```{r add_mouseMine_and_Open_Targets_data_QTL}
queryMouseMine(project_dir = "qtl_analysis")
queryOpenTargets(project_dir = "qtl_analysis") 
```

### Examine Package Structure Before Creating Outputs

```{r gene_examine_package_QTL, message=T}
listPackRatTables(project_dir = "qtl_analysis", full_info = TRUE)

```

### Generate Locus Plot

```{r generate_QTL_plot}
generateLocusZoomPlot(
     region_id="region_1",
     project_dir="qtl_analysis",
     scan_table="QTL_results",
     signal_table="founder_info",
     width=10,
     height=6,
     threshold=4,
     layout_ratios = c(manhattan = 0.35, signal = 0.40, genes = 0.25))

```

### Generate Excel Sheet

Additional sheets can be created from any boolean (True/False) argument.
Boolean statements in R can be strung together with & (and) or | (or).
We recommend examining the package structure above with full_info = TRUE to identify column names.

For Region-based packages, one likely does not wish to include all region-based information in the output (for example, the LoD scores from the genome scan itself).  The exclude_tables argument will allow you to exclude these from your output

Note: beware of capitalization.  For some reason mouseMine does not capitalize any terms other than gene names, while OpenTargets is somewhat inconsistent.  If needed, more complex arguments e.g. grepl("[c|C]ardiac",mm_phenotype) & padj <.001 may be used.

Note 2:  If two supplemental tables share a column name, the code will append the table's abbreviation to the column name: (e.g. if the table abbreviation is 'ab' and the column is 'value', it will become 'ab_value').  It is also possible to set the prefix_mode argument below to "abbreviated", in which case all columns from tables with an abbreviation will have it appended by default.

```{r generate_region_excels_QTL}
# Peak-centric analysis workbook
makeGeneSheet(
  format = "excel",
  output_file = "qtl_analysis.xlsx",
  prefix_mode = "abbreviated",
  split_by = "criteria",
  split_criteria = list(
    "All_Results" = "TRUE",
    "Chr 3 Peak" = "peak_id == 'Significant_1'",
    "Mouse_Phenotype" = "grepl('[c|C]ardiac',mm_phenotype) | grepl('[h|H]eart',mm_phenotype) ",
    "Sig OTC" = "(otc_score>1 | otc_score< -1) & peak_id == 'Significant_1'"
  ),
  include_supplementary =  TRUE,
  exclude_tables=c("founder_info","QTL_results"),
  project_dir = "qtl_analysis"
)

```

## ATACseq Project

### Initialize Project with Genomic Regions - ATACseq

```{r region_mode_init_ATAC, message=TRUE}
# Prepare example ATAC-seq peaks
set.seed(123)
chromosomes <- c(as.character(1:19), "X", "Y")

atac_peaks <- data.table(
  chr = sample(chromosomes, 200, replace = TRUE,
               prob = c(rep(0.06, 19), 0.07, 0.07)),
  start = sample(1:150000000, 200),
  peak_id = paste0("peak_", 1:200),
  fold_enrichment = round(runif(200, 2, 50), 2),
  qvalue = 10^(-runif(200, 1, 20))
)

# Calculate end positions (typical ATAC peak width: 200-1000bp)
atac_peaks[, end := start + sample(200:1000, .N, replace = TRUE)]
atac_peaks[, summit := start + round((end - start) / 2)]
atac_peaks[, peak_type := sample(c("promoter", "enhancer", "intergenic"),
                                 .N, replace = TRUE,
                                 prob = c(0.3, 0.4, 0.3))]

# Initialize region-based project
initPackRat(
  data = atac_peaks,
  mode = "region",
  species = "mouse",
  genome = "mm39",
  project_dir = "atacseq_analysis",
  force = TRUE
)
```

### Add Chromatin Annotations

#### Chromatin States

```{r add_chromatin}
# Load processed regions
regions <- fread("atacseq_analysis/.locusPackRat/input/regions.csv")

# Add chromatin state annotations
chromatin_states <- data.table(
  chr = regions$chr,
  start = regions$start,
  end = regions$end,
  state = sample(c("Active_Promoter", "Strong_Enhancer", "Weak_Enhancer",
                  "Poised_Promoter", "Repressed", "Heterochromatin", "Transcribed"),
                size = nrow(regions), replace = TRUE,
                prob = c(0.15, 0.2, 0.15, 0.1, 0.1, 0.1, 0.2)),
  cell_type = sample(c("ES_cells", "Neural", "Hepatocyte", "T_cells"),
                    size = nrow(regions), replace = TRUE)
)

addRatTable(
  data = chromatin_states,
  table_name = "chromatin_states",
  abbreviation = "cs",
  link_type = "region",
  link_by = "chr,start,end",
  project_dir = "atacseq_analysis"
)
```

#### Transcription Factor Motifs

```{r add_tf_motifs}
# Add TF binding motif data
tf_list <- c("CTCF", "NFKB", "AP1", "ETS1", "GATA1", "SOX2", "OCT4", "NANOG",
            "MYC", "MAX", "STAT3", "CREB", "SP1", "YY1", "REST")

# Create random starts
random_starts <- sample(regions$start, 120, replace = TRUE)

motif_data <- data.table(
  chr = sample(regions$chr, 120, replace = TRUE),
  start = random_starts,
  end = random_starts + sample(50:200, 120, replace = TRUE),
  tf_name = sample(tf_list, 120, replace = TRUE),
  motif_score = round(runif(120, 6, 15), 2),
  p_value = 10^(-runif(120, 2, 8))
)

# Add additional motifs for some peaks
additional_motifs <- motif_data[sample(1:nrow(motif_data), 30)]
additional_motifs[, tf_name := sample(tf_list, .N, replace = TRUE)]
motif_data <- rbind(motif_data, additional_motifs)

addRatTable(
  data = motif_data,
  table_name = "tf_motifs",
  abbreviation = "tf_motifs",
  link_type = "region",
  link_by = "chr,start,end",
  project_dir = "atacseq_analysis"
)
```

#### Differential Accessibility

```{r add_differential}
# Add differential accessibility between conditions
diff_access <- data.table(
  chr = regions$chr,
  start = regions$start,
  end = regions$end,
  condition_A_signal = round(abs(rnorm(nrow(regions), 100, 50)), 1),
  condition_B_signal = round(abs(rnorm(nrow(regions), 100, 50)), 1),
  log2FC = round(rnorm(nrow(regions), 0, 1.5), 2),
  padj = 10^(-runif(nrow(regions), 0, 8)),
  higher_in = ifelse(rnorm(nrow(regions)) > 0, "condition_B", "condition_A"),
  significant = runif(nrow(regions)) < 0.3
)

addRatTable(
  data = diff_access,
  table_name = "differential_accessibility",
  abbreviation = "da",
  link_type = "region",
  link_by = "chr,start,end",
  project_dir = "atacseq_analysis"
)
```

#### Add mouseMine and Open Targets data

```{r add_mouseMine_and_Open_Targets_data_ATAC}
queryMouseMine(project_dir = "atacseq_analysis")
queryOpenTargets(project_dir = "atacseq_analysis") 
```

### Examine Package Structure Before Creating Output

```{r gene_examine_package_ATAC, message=T}
listPackRatTables(project_dir = "atacseq_analysis", full_info = TRUE)

```

### Generate ATACseq Region Analysis Workbooks

Additional sheets can be created from any boolean (True/False) argument.
Boolean statements in R can be strung together with & (and) or | (or).
We recommend examining the package structure above with full_info = TRUE to identify column names.

For Region-based packages, one likely does not wish to include all region-based information in the output (for example, the LoD scores from the genome scan itself).  See below how to use include_supplementary to include only the tables you desire

Note: beware of capitalization.  For some reason mouseMine does not capitalize any terms other than gene names, while OpenTargets is somewhat inconsistent.  If needed, more complex arguments e.g. grepl("[c|C]ardiac",mm_phenotype) & padj <.001 may be used.

Note 2:  If two supplemental tables share a column name, the code will append the table's abbreviation to the column name: (e.g. if the table abbreviation is 'ab' and the column is 'value', it will become 'ab_value').  It is also possible to set the prefix_mode argument below to "abbreviated", in which case all columns from tables with an abbreviation will have it appended by default.

```{r generate_region_excels}
# Peak-centric analysis workbook
makeGeneSheet(
  format = "excel",
  output_file = "atacseq_peak_analysis.xlsx",
  split_by = "criteria",
  split_criteria = list(
    "All_Peaks" = "TRUE",
    "High_Enrichment" = "fold_enrichment > 20",
    "Promoter_Peaks" = "peak_type == 'promoter'",
    "Enhancer_Peaks" = "peak_type == 'enhancer'",
    "Active_Chromatin" = "state %in% c('Active_Promoter', 'Strong_Enhancer')",
    "Differential" = "significant == TRUE & abs(log2FC) > 1"
  ),
  include_supplementary = TRUE,
  project_dir = "atacseq_analysis"
)

# TF-centric analysis workbook
makeGeneSheet(
  format = "excel",
  output_file = "atacseq_tf_binding.xlsx",
  split_by = "criteria",
  split_criteria = list(
    "CTCF_Binding" = "tf_name == 'CTCF'",
    "Pioneer_TFs" = "tf_name %in% c('SOX2', 'OCT4', 'NANOG')",
    "Immune_TFs" = "tf_name %in% c('NFKB', 'AP1', 'STAT3')",
    "High_Motif_Score" = "motif_score > 12"
  ),
  include_supplementary = c("tf_motifs", "chromatin_states"),
  project_dir = "atacseq_analysis"
)
```

# Working Across Species and Genome Builds

## Human Gene Analysis with Mouse Orthologs

The package automatically handles orthology mapping between human and mouse using the built-in coordinate files.


```{r crossspecies_analysis, message=TRUE}
# Define human cancer genes
cancer_genes <- c("BRCA1", "BRCA2", "TP53", "EGFR", "KRAS", "MYC",
                 "PTEN", "RB1", "VHL", "APC", "ATM", "CDKN2A",
                 "MLH1", "MSH2", "BRAF", "PIK3CA", "ERBB2", "ALK")

human_genes <- data.table(
  gene_symbol = cancer_genes,
  cancer_type = c(rep("breast", 2), "multiple", "lung", "colorectal", "multiple",
                  "brain", "retinoblastoma", "kidney", "colorectal", "breast", "melanoma",
                  rep("colorectal", 2), "melanoma", "multiple", "breast", "lung"),
  mutation_frequency = round(runif(length(cancer_genes), 5, 80), 1)
)

# Initialize with automatic orthology mapping
initPackRat(
  data = human_genes,
  mode = "gene",
  species = "human",
  genome = "hg38",
  project_dir = "cancer_genes",
  force = TRUE
)

# Add clinical significance data
clinical_data <- data.table(
  gene_symbol = cancer_genes,
  pathogenicity = sample(c("pathogenic", "likely_pathogenic", "VUS"),
                        length(cancer_genes), replace = TRUE,
                        prob = c(0.5, 0.3, 0.2)),
  FDA_approved_drug = sample(c(TRUE, FALSE), length(cancer_genes),
                            replace = TRUE, prob = c(0.3, 0.7)),
  clinical_trials = sample(0:10, length(cancer_genes), replace = TRUE)
)

addRatTable(
  data = clinical_data,
  table_name = "clinical_significance",
  link_type = "gene",
  link_by = "gene_symbol",
  project_dir = "cancer_genes"
)

# Generate cross-species analysis workbook
makeGeneSheet(
  format = "excel",
  output_file = "cancer_genes_analysis.xlsx",
  split_by = "criteria",
  split_criteria = list(
    "High_Frequency" = "mutation_frequency > 30",
    "FDA_Targets" = "FDA_approved_drug == TRUE",
    "Breast_Cancer" = "cancer_type == 'breast'",
    "With_Mouse_Ortholog" = "!is.na(mouse_gene_symbol)"
  ),
  include_supplementary = TRUE,
  project_dir = "cancer_genes"
)
```

## Converting Between Genome Coordinates

In some cases, users may have data that use older genome builds (e.g. mm9) and wish to incorporate them into a project based around a newer build (e.g. mm39).

Although LocusPackRat does not have a built-in function to accomodate other builds, the liftOver package, paired with chain files drawn from the UCSC genome browser, can easily do these conversions.

```{r liftOver, message=F, eval=requireNamespace("liftOver", quietly=TRUE)}
library(liftOver)
mm9_ncRNA = read.csv(system.file("extdata", "sample_mm9_ncRNA.csv", package = "locusPackRat"))
chain = import.chain( system.file("extdata", "mm9ToMm39.over.chain", package = "locusPackRat"))

#convert table to GRanges object.  Note that the chain files normally want
#chromosomes in 'chr1' format vs '1' format.  So you will need to convert.
mm9_ncRNA$chr=paste0("chr",mm9_ncRNA$chr)
mm9_ncRNA=makeGRangesFromDataFrame(mm9_ncRNA,keep.extra.columns = T)

mm39_ncRNA=unlist(liftOver(mm9_ncRNA,chain = chain))
mm39_ncRNA=as.data.frame(mm39_ncRNA)
mm39_ncRNA$seqnames=gsub("chr","",mm39_ncRNA$seqnames)

#Your data are now ready to be added to a locusPackRat project


```


# Working with Projects

## List Available Tables

```{r list_tables, message=T}
# View supplementary tables in each project
listPackRatTables("rnaseq_analysis", full_info = TRUE)
listPackRatTables("atacseq_analysis", full_info = FALSE)
```

## Remove a Table

```{r remove_table}

removeRatTable("chromatin_states", project_dir = "atacseq_analysis")

```


## Project Configuration

Each project contains a `config.json` file with project metadata:

```{r show_config, eval=FALSE}
# Load project configuration
config <- jsonlite::read_json("rnaseq_analysis/.locusPackRat/config.json")
str(config)
```

## Output Options

### CSV Format
```{r csv_output, eval=FALSE}
makeGeneSheet(
  format = "csv",
  output_file = "results.csv",
  include_supplementary = TRUE,
  project_dir = "rnaseq_analysis"
)
```

### Filtered Output
```{r filtered_output, eval=FALSE}
makeGeneSheet(
  format = "csv",
  output_file = "significant_genes.csv",
  filter_expr = "padj < 0.05 & abs(log2FC) > 1",
  include_supplementary = TRUE,
  project_dir = "rnaseq_analysis"
)
```

### Selective Table Inclusion
```{r selective_tables, eval=FALSE}
makeGeneSheet(
  format = "excel",
  output_file = "selected_data.xlsx",
  include_supplementary = c("tissue_expression", "pathway_annotations"),
  project_dir = "rnaseq_analysis"
)
```

# Converting Genome Builds

All data added to a locusPackRat project must use the same genome build (e.g., mm39, hg38). If you have data in a different build, use `rtracklayer::liftOver()` to convert coordinates before adding them to the project.

```{r liftover_example, eval=FALSE}
# Example: Convert mm10 coordinates to mm39
library(rtracklayer)
library(GenomicRanges)

# Load a chain file for the conversion
# Chain files are available from UCSC:
# https://hgdownload.cse.ucsc.edu/goldenpath/mm10/liftOver/mm10ToMm39.over.chain.gz
chain <- import.chain("mm10ToMm39.over.chain")

# Create a GRanges object from your data
# Suppose you have eQTL data in mm10 coordinates
eqtl_mm10 <- data.table(
  chr = c("chr3", "chr3", "chr3"),
  start = c(131000000, 135000000, 142000000),
  end = c(131001000, 135001000, 142001000),
  gene = c("Cisd2", "Manba", "Pdlim5"),
  lod = c(5.2, 4.8, 3.1)
)

gr_mm10 <- GRanges(
  seqnames = eqtl_mm10$chr,
  ranges = IRanges(start = eqtl_mm10$start, end = eqtl_mm10$end),
  gene = eqtl_mm10$gene,
  lod = eqtl_mm10$lod
)

# Perform liftOver
gr_mm39 <- liftOver(gr_mm10, chain)

# Convert back to data.table — liftOver returns a GRangesList
# (one element per input range, since some ranges may map to multiple locations)
gr_mm39_unlisted <- unlist(gr_mm39)

eqtl_mm39 <- data.table(
  chr = as.character(seqnames(gr_mm39_unlisted)),
  start = start(gr_mm39_unlisted),
  end = end(gr_mm39_unlisted),
  gene = gr_mm39_unlisted$gene,
  lod = gr_mm39_unlisted$lod
)

# Remove "chr" prefix if your project uses UCSC-style names without it
eqtl_mm39[, chr := gsub("^chr", "", chr)]

# Now add to your mm39 locusPackRat project
addRatTable(
  data = eqtl_mm39,
  table_name = "eqtl_lifted",
  abbreviation = "eql",
  link_type = "region",
  link_by = "chr,start,end",
  project_dir = "qtl_analysis"
)
```

# Session Information

```{r session}
sessionInfo()
```


