---
title: "New UX Features in locusPackRat v0.7"
author: "Brian Gural"
date: "`r Sys.Date()`"
output:
    rmarkdown::html_vignette:
        toc: true
        toc_depth: 2
vignette: >
  %\VignetteIndexEntry{New UX Features in locusPackRat v0.7}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  warning = FALSE
)
```

# What's new

locusPackRat v0.7 introduces several quality-of-life improvements:

1. **`lpr_*` aliases** — every function now has a short, tab-completable alias
2. **`lpr_check()`** — a diagnostic command that summarizes your project and flags issues
3. **Lighter install** — `plotgardener` and `httr` are now optional (only needed when you use plotting or API queries)
4. **Cross-platform zip** — `buildPacket()` works without a system `zip` command if the `zip` R package is installed

# The `lpr_*` aliases

All original function names still work. The aliases just give you a common `lpr_` prefix
so you can type `lpr_` and Tab to see everything:

| Original | Alias |
|---|---|
| `initPackRat()` | `lpr_init()` |
| `addRatTable()` | `lpr_add_table()` |
| `removeRatTable()` | `lpr_remove_table()` |
| `listPackRatTables()` | `lpr_list_tables()` |
| `makeGeneSheet()` | `lpr_export()` |
| `buildPacket()` | `lpr_packet()` |
| `generateLocusZoomPlot()` | `lpr_zoom()` |
| `queryMouseMine()` | `lpr_query_mousemine()` |
| `queryOpenTargets()` | `lpr_query_ot()` |
| `queryOpenTargetsQTL()` | `lpr_query_ot_qtl()` |
| `filterGenes()` | `lpr_filter()` |
| `makeFilter()` | `lpr_make_filter()` |

They are identical — `lpr_init` **is** `initPackRat`, not a wrapper:

```{r alias-identity}
library(locusPackRat)
identical(lpr_init, initPackRat)
identical(lpr_export, makeGeneSheet)
```

# Quick workflow using aliases

```{r demo-workflow, message=TRUE}
# Create a small demo project
demo_dir <- file.path(tempdir(), "lpr_demo")

lpr_init(
  data = data.frame(
    gene_symbol = c("Myh7", "Tnnt2", "Nppa", "Nppb", "Atp2a2"),
    log2FC      = c(2.1, -0.8, 3.5, 2.9, -1.2)
  ),
  mode = "gene",
  species = "mouse",
  genome = "mm39",
  project_dir = demo_dir,
  force = TRUE
)
```

## Adding supplementary data

```{r add-table, message=TRUE}
# Some made-up expression data
expr_data <- data.frame(
  gene_symbol = c("Myh7", "Tnnt2", "Nppa", "Nppb", "Atp2a2"),
  tissue      = c("heart", "heart", "heart", "heart", "heart"),
  tpm         = c(150.2, 89.4, 210.5, 180.3, 95.1)
)

lpr_add_table(
  expr_data,
  table_name = "tissue_expression",
  link_type = "gene",
  link_by = "gene_symbol",
  abbreviation = "te",
  project_dir = demo_dir
)
```

## Listing tables

```{r list-tables, message=TRUE}
lpr_list_tables(demo_dir)
```

# Project diagnostics with `lpr_check()`

`lpr_check()` gives you a quick summary of a project's state — metadata, tables,
outputs, and optional dependency availability:

```{r check-demo, message=TRUE}
lpr_check(demo_dir)
```

You can also capture the result for programmatic use:

```{r check-result}
status <- lpr_check(demo_dir, verbose = FALSE)
status$valid
status$config$species
status$config$genome
```

## Running without a project

If you point `lpr_check()` at a directory with no project, it tells you what to do:

```{r check-no-project, message=TRUE}
lpr_check(tempdir())
```

# Exporting a gene sheet

```{r export-csv, message=TRUE}
result <- lpr_export(
  format = "csv",
  output_file = "demo_sheet.csv",
  project_dir = demo_dir
)

# The result is the gene sheet as a data.table (returned invisibly)
head(result)
```

## Checking output staleness

After exporting, `lpr_check()` tracks whether outputs are up to date. If you add
more data after exporting, it will flag the output as stale:

```{r staleness, message=TRUE}
# Add another table after the export
Sys.sleep(1)  # ensure file timestamp differs
new_data <- data.frame(
  gene_symbol = c("Myh7", "Tnnt2", "Nppa", "Nppb", "Atp2a2"),
  pathway     = c("sarcomere", "sarcomere", "natriuretic", "natriuretic", "calcium")
)

lpr_add_table(
  new_data,
  table_name = "pathways",
  link_type = "gene",
  link_by = "gene_symbol",
  abbreviation = "pw",
  project_dir = demo_dir
)

# Now lpr_check will flag that the output is older than supplementary data
lpr_check(demo_dir)
```

# Optional dependencies

`plotgardener` and `httr` are no longer required at install time. They are only needed
when you call the functions that use them:

- **`lpr_zoom()`** needs `plotgardener` + a TxDb package for your genome
- **`lpr_query_mousemine()`**, **`lpr_query_ot()`**, **`lpr_query_ot_qtl()`** need `httr`

If a dependency is missing, you get a clear error with install instructions rather
than a cryptic namespace failure.

`lpr_check()` reports which optional dependencies are available in its dependency
checks section.

```{r cleanup, include=FALSE}
unlink(demo_dir, recursive = TRUE)
```
