---
title: "Region-Level QTL Analysis with Open Targets Integration"
author: "Brian Gural, Todd Kimball, Anh Luu, Christoph D. Rau"
date: "`r Sys.Date()`"
output:
    rmarkdown::html_vignette:
        toc: true
        toc_depth: 3
vignette: >
  %\VignetteIndexEntry{Region-Level QTL Analysis with Open Targets Integration}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.width = 7,
  fig.height = 5,
  warning = FALSE,
  message = FALSE
)

# Check for API availability
# Skip on CRAN (R CMD check sets _R_CHECK_PACKAGE_NAME_) or when API unreachable
on_cran <- nzchar(Sys.getenv("_R_CHECK_PACKAGE_NAME_"))

# Check if we can reach the API endpoint (any response means it's reachable)
api_available <- requireNamespace("httr", quietly = TRUE) && {
  resp <- try(httr::GET("https://api.platform.opentargets.org/api/v4/graphql",
                        httr::timeout(10)), silent = TRUE)
  !inherits(resp, "try-error")
}

skip_api <- on_cran || !api_available

# Check for Bioconductor annotation packages needed for locus zoom plots
has_bioc_annot <- requireNamespace("TxDb.Hsapiens.UCSC.hg38.knownGene", quietly = TRUE) &&
                  requireNamespace("org.Hs.eg.db", quietly = TRUE) &&
                  requireNamespace("GenomicFeatures", quietly = TRUE)
can_plot <- !skip_api && has_bioc_annot
```

# Overview

This vignette demonstrates a region-level QTL analysis workflow using `locusPackRat`. We focus on three human genomic regions (~2 MB each) that have documented eQTL and pQTL data in Open Targets:
1. **ZNF586 locus** (chr19): Region with extensive eQTL data
2. **APIP locus** (chr11): Region with expression QTL associations
3. **CLEC19A locus** (chr16): Region with protein QTL data

We showcase the Open Targets Platform integration functions:

- `queryOpenTargets()`: Disease associations, genetic constraints, and tractability
- `queryOpenTargetsQTL()`: eQTL and pQTL data from fine-mapped credible sets

# Setup

```{r libraries}
library(locusPackRat)
library(data.table)
```

# Define QTL Regions

We define three regions of approximately 2 MB each, selected because they have documented eQTL and pQTL data in the Open Targets Platform.

```{r define_regions}
qtl_regions <- data.table(
  chr = c("19", "11", "16"),
  start = c(56769655, 33875225, 18285731),
  end = c(58769655, 35875225, 20285731),
  region_id = c("ZNF586_locus", "APIP_locus", "CLEC19A_locus"),
  peak_lod = c(8.2, 6.5, 7.1),
  trait = c("gene_expression", "gene_expression", "protein_levels")
)

# Display region summary
qtl_regions[, .(
  region_id,
  chr,
  size_kb = round((end - start) / 1000),
  trait
)]
```

# Initialize Project

Initialize a region-mode project with human hg38 coordinates.

```{r init_project}
initPackRat(
  data = qtl_regions,
  mode = "region",
  species = "human",
  genome = "hg38",
  project_dir = "qtl_opentargets_demo",
  force = TRUE
)
```

# Add Simulated LOD Data

For demonstration purposes, we generate simulated QTL scan data with Gaussian peaks centered in each region. In a real analysis, this would be replaced with actual scan results from qtl2 or similar software.

```{r generate_lod_helper}
# Helper function to generate realistic LOD curves
generate_lod_data <- function(regions, points_per_region = 500,
                              noise_sd = 0.3) {
  lod_list <- lapply(seq_len(nrow(regions)), function(i) {
    region <- regions[i, ]
    positions <- seq(region$start, region$end, length.out = points_per_region)

    # Create Gaussian peak centered at midpoint
    peak_pos <- (region$start + region$end) / 2
    peak_width <- (region$end - region$start) / 6

    # Base LOD from Gaussian curve + noise
    base_lod <- region$peak_lod *
      exp(-((positions - peak_pos)^2) / (2 * peak_width^2))
    lod_values <- pmax(0, base_lod + rnorm(length(positions), 0, noise_sd))

    data.table(
      chr = region$chr,
      pos = as.integer(positions),
      lod = round(lod_values, 3),
      region_id = region$region_id,
      trait = region$trait
    )
  })

  rbindlist(lod_list)
}

# Generate LOD data
set.seed(42)
lod_scan <- generate_lod_data(qtl_regions)

# Preview
head(lod_scan)
```

```{r add_lod_table}
# Add scan data to project
addRatTable(
  data = lod_scan,
  table_name = "qtl_scan",
  link_type = "point",
  project_dir = "qtl_opentargets_demo"
)
```

# Query Open Targets Platform

The Open Targets Platform provides comprehensive gene annotations including disease associations, genetic constraints (gnomAD), and drug tractability assessments.

```{r query_open_targets, eval=!skip_api}
queryOpenTargets(
  project_dir = "qtl_opentargets_demo",
  data_types = c("diseases", "constraints", "tractability"),
  disease_limit = 100,  # Limit for vignette performance
  verbose = TRUE
)
```

```{r skip_message_ot, eval=skip_api, echo=FALSE}
message("Skipping Open Targets Platform query (no internet or running on CRAN)")
```

# Query Open Targets QTL Data

The `queryOpenTargetsQTL()` function retrieves fine-mapped eQTL and pQTL data from credible sets, including locus-to-gene (L2G) prediction scores. The function queries by genomic region, making it suitable for both gene-mode and region-mode projects.

```{r query_open_targets_qtl, eval=!skip_api, message=TRUE}
qtl_results <- queryOpenTargetsQTL(
  project_dir = "qtl_opentargets_demo",
  study_types = c("eqtl", "pqtl"),
  include_l2g = TRUE,
  min_l2g_score = 0.5,
  verbose = TRUE
)
```

```{r skip_message_qtl, eval=skip_api, echo=FALSE}
message("Skipping Open Targets QTL query (no internet or running on CRAN)")
```

## QTL Studies Table

The studies table contains metadata about each QTL study, including tissue/cell type (biosample), sample size, and the target gene.

```{r qtl_studies_preview, eval=!skip_api, echo=TRUE, results='markup'}
if (!is.null(qtl_results$studies) && nrow(qtl_results$studies) > 0) {
  studies_dt <- qtl_results$studies
  cat("QTL Studies table:", nrow(studies_dt), "rows\n\n")

  # Select key columns for display
  display_cols <- intersect(
    c("study_id", "study_type", "biosample_name",
      "n_samples", "target_gene_symbol"),
    names(studies_dt)
  )

  cat("Head of studies table:\n")
  print(head(studies_dt[, ..display_cols], 5))

  cat("\nTail of studies table:\n")
  print(tail(studies_dt[, ..display_cols], 5))
} else {
  cat("No QTL studies found for the specified regions.\n")
}
```

## QTL Credible Sets Table

The credible sets table contains fine-mapped QTL associations with lead variants, effect sizes, and L2G scores linking variants to their most likely causal genes.

```{r qtl_credsets_preview, eval=!skip_api, echo=TRUE, results='markup'}
credsets_dt <- qtl_results$credible_sets
if (!is.null(credsets_dt) && nrow(credsets_dt) > 0) {
  cat("QTL Credible Sets table:", nrow(credsets_dt), "rows\n\n")

  # Select key columns for display (L2G columns omitted -- see note below)
  display_cols <- intersect(
    c("study_id", "study_type", "chromosome", "position", "rsid",
      "beta", "pvalue_mantissa", "pvalue_exponent"),
    names(credsets_dt)
  )

  cat("Head of credible sets table:\n")
  print(head(credsets_dt[, ..display_cols], 5))

  cat("\nTail of credible sets table:\n")
  print(tail(credsets_dt[, ..display_cols], 5))

  # Summary statistics
  cat("\n\nSummary by study type:\n")
  print(credsets_dt[, .N, by = study_type])

  cat("\nNote: L2G (locus-to-gene) prediction scores are primarily available",
      "for GWAS credible sets.\n",
      "molQTL studies (eQTL/pQTL) typically return empty L2G predictions",
      "because the\n",
      "QTL target gene is already known from the study design.",
      "The l2g_gene_symbol and\n",
      "l2g_score columns are retained in the stored data table",
      "for use with GWAS queries.\n")
} else {
  cat("No QTL credible sets found for the specified regions.\n")
}
```

# Examine Project Structure

Review all tables added to the project.

```{r list_tables, message=TRUE}
listPackRatTables(project_dir = "qtl_opentargets_demo", full_info = TRUE)
```

# Generate Locus Zoom Plots

Create publication-quality locus zoom plots for each region. These plots show the QTL scan results alongside gene annotations.

**Note:** The gene annotation plots require Bioconductor annotation packages. Install them with:
```r
BiocManager::install(c("TxDb.Hsapiens.UCSC.hg38.knownGene", "org.Hs.eg.db"))
```

```{r locus_zoom_znf586, eval=can_plot}
# ZNF586 locus - expression QTL region
generateLocusZoomPlot(
  region_id = "ZNF586_locus",
  project_dir = "qtl_opentargets_demo",
  scan_table = "qtl_scan",
  width = 8,
  height = 5,
  highlight_genes = c("ZNF586"),
  threshold = 5,
  output_file = "ZNF586_locus_zoom.png"
)
knitr::include_graphics(
  file.path("qtl_opentargets_demo", ".locusPackRat", "output", "ZNF586_locus_zoom.png")
)
```

```{r locus_zoom_apip, eval=can_plot}
# APIP locus - expression QTL region
generateLocusZoomPlot(
  region_id = "APIP_locus",
  project_dir = "qtl_opentargets_demo",
  scan_table = "qtl_scan",
  width = 8,
  height = 5,
  highlight_genes = c("APIP"),
  threshold = 5,
  output_file = "APIP_locus_zoom.png"
)
knitr::include_graphics(
  file.path("qtl_opentargets_demo", ".locusPackRat", "output", "APIP_locus_zoom.png")
)
```

```{r locus_zoom_clec19a, eval=can_plot}
# CLEC19A locus - protein QTL region
generateLocusZoomPlot(
  region_id = "CLEC19A_locus",
  project_dir = "qtl_opentargets_demo",
  scan_table = "qtl_scan",
  width = 8,
  height = 5,
  highlight_genes = c("CLEC19A"),
  threshold = 5,
  output_file = "CLEC19A_locus_zoom.png"
)
knitr::include_graphics(
  file.path("qtl_opentargets_demo", ".locusPackRat", "output", "CLEC19A_locus_zoom.png")
)
```

# Export Results

Generate a multi-sheet Excel workbook with filtered views of the integrated data.

```{r export_excel, eval=FALSE}
makeGeneSheet(
  format = "excel",
  output_file = "qtl_opentargets_results.xlsx",
  split_by = "criteria",
  prefix_mode = "collision",
  split_criteria = list(
    "All_Genes" = "TRUE",
    "High_LOD" = "peak_lod > 7",
    "Druggable" = "!is.na(ott_label)",
    "Constrained" = "otc_constraintType == 'lof' & otc_oe < 0.35"
  ),
  project_dir = "qtl_opentargets_demo"
)
```

# Notes and Caveats

## API Rate Limits

The Open Targets API has rate limits. For large gene sets, consider:

- Using the `limit` parameter for testing
- Running queries during off-peak hours
- Caching results locally

## Species Handling

`queryOpenTargets()` and `queryOpenTargetsQTL()` query human data. For mouse projects, genes are automatically mapped to human orthologs before querying.

## Offline Usage

When running without internet access (or on CRAN), the API query chunks are skipped. Pre-computed results can be loaded from local files.

## L2G Scores

Locus-to-Gene (L2G) scores range from 0 to 1, where higher scores indicate stronger evidence linking a variant to a gene. The default threshold of 0.5 provides good specificity.

# Session Info

```{r session_info}
tryCatch(sessionInfo(), error = function(e) message("sessionInfo() unavailable"))
```

# Cleanup

```{r cleanup, include=FALSE}
# Deferred so PNGs still exist when pandoc embeds them.
reg.finalizer(
  environment(),
  function(e) unlink("qtl_opentargets_demo", recursive = TRUE),
  onexit = TRUE
)
```
