---
title: "Single-Cell RNA-seq Integration for Cell-Type-Specific Candidate Evaluation"
author: "Brian Gural, Todd Kimball, Anh Luu, Christoph D. Rau"
date: "`r Sys.Date()`"
output:
    rmarkdown::html_vignette:
        toc: true
        toc_depth: 3
vignette: >
  %\VignetteIndexEntry{Single-Cell RNA-seq Integration for Cell-Type-Specific Candidate Evaluation}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.width = 7,
  fig.height = 5,
  eval = requireNamespace("Seurat", quietly = TRUE) &&
         requireNamespace("ggplot2", quietly = TRUE)
)
```

# Overview

Bulk-tissue transcriptomic data, which form a common evidence layer in GWAS
candidate gene prioritization, can obscure effects that are specific to
particular cell types. A gene that appears broadly expressed in bulk tissue
may in fact be restricted to a small cell population, and conversely, a gene
with moderate bulk expression may be highly enriched in the disease-relevant
cell type.

This vignette demonstrates how to use single-cell RNA-seq (scRNA-seq) data
to evaluate candidate genes identified by `locusPackRat` for cell-type-specific
expression. Specifically, we show how to:

1. Load a reference scRNA-seq dataset
2. Score candidate gene sets using Seurat's `AddModuleScore()`
3. Visualize module scores across cell types
4. Use the results to refine candidate gene prioritization

A bundled mini Seurat object (200 cells, 4 cardiac cell types) is used here
for demonstration. For real analyses, substitute your own scRNA-seq reference
(see the Tabula Muris example at the end of this vignette).

# Prerequisites

```{r libraries}
# Core packages
library(locusPackRat)
library(data.table)

# scRNA-seq analysis
library(Seurat)     # install from CRAN: install.packages("Seurat")
library(ggplot2)
```

# Step 1: Create a locusPackRat Project

First, we create a temporary locusPackRat project with candidate genes from a
cardiac hypertrophy study. These five genes will be evaluated for cell-type
specificity.

```{r create_project}
# Define candidate genes from a cardiac QTL study
cardiac_candidates <- c("Cisd2", "Pdlim5", "Manba", "Fhod3", "Myh7")

# Create a temporary project directory
project_dir <- file.path(tempdir(), "sc_demo_project")

# Initialize a locusPackRat project with these genes
gene_input <- data.frame(gene_symbol = cardiac_candidates)
initPackRat(
  data = gene_input,
  mode = "gene",
  species = "mouse",
  genome = "mm39",
  project_dir = project_dir,
  force = TRUE
)

# Show the initialized gene table
master_genes <- fread(file.path(project_dir, ".locusPackRat", "input", "genes.csv"))
print(master_genes[, .(gene_symbol, chr, start, end)])
```

# Step 2: Load a Reference scRNA-seq Dataset

We load the bundled mini Seurat object, which contains 200 cells across four
cardiac cell types (cardiomyocyte, fibroblast, endothelial, immune) with
pre-computed PCA and UMAP embeddings.

```{r load_data}
# Load the bundled mini Seurat object
seurat_path <- system.file("extdata", "mini_heart_seurat.rds",
                           package = "locusPackRat")
# Fallback for devtools/development builds where inst/ isn't installed yet
if (!nzchar(seurat_path)) {
  seurat_path <- file.path("..", "inst", "extdata", "mini_heart_seurat.rds")
}

if (!file.exists(seurat_path)) {
  message("mini_heart_seurat.rds not found -- skipping remaining chunks.")
  knitr::opts_chunk$set(eval = FALSE)
} else {
  heart_seurat <- readRDS(seurat_path)

  # Inspect the object
  heart_seurat

  # Cell type composition
  table(heart_seurat$cell_type)
}
```

For a real analysis, you would load a full scRNA-seq reference such as
Tabula Muris. The code below is provided as a reference but not executed:

```{r load_tabula_muris, eval = FALSE}
# Download Tabula Muris heart data (10X Genomics)
# Data available from: https://tabula-muris.ds.czbiohub.org/
heart_data <- Read10X(data.dir = "path/to/tabula_muris/Heart-10X_P7_4/")
heart_seurat <- CreateSeuratObject(counts = heart_data, project = "TabulaMuris_Heart",
                                   min.cells = 3, min.features = 200)
heart_seurat <- NormalizeData(heart_seurat)
heart_seurat <- FindVariableFeatures(heart_seurat, nfeatures = 2000)
heart_seurat <- ScaleData(heart_seurat)
heart_seurat <- RunPCA(heart_seurat)
heart_seurat <- FindNeighbors(heart_seurat, dims = 1:20)
heart_seurat <- FindClusters(heart_seurat, resolution = 0.5)
heart_seurat <- RunUMAP(heart_seurat, dims = 1:20)
```

# Step 3: Score Candidate Genes with AddModuleScore

Seurat's `AddModuleScore()` calculates per-cell enrichment scores for a set
of genes, controlling for expression level and gene set size. This allows us
to ask: "Are our candidate genes collectively enriched in specific cell types?"

```{r module_scores}
# Only score genes that are present in the dataset
candidates_in_data <- intersect(cardiac_candidates, rownames(heart_seurat))
cat("Candidates found in dataset:", paste(candidates_in_data, collapse = ", "), "\n")

# Define gene lists for module scoring
candidate_list <- list(cardiac_candidates = candidates_in_data)

# Calculate module scores
# For small gene panels (e.g., bundled demo data), reduce ctrl and nbin
# so that control gene sampling does not exceed available genes per bin
n_features <- nrow(heart_seurat)
use_ctrl <- if (n_features < 200) 5 else 100
use_nbin <- if (n_features < 200) 5 else 24
heart_seurat <- AddModuleScore(
  object = heart_seurat,
  features = candidate_list,
  ctrl = use_ctrl,
  nbin = use_nbin,
  name = "candidate_score"
)

# The score is added as a metadata column: "candidate_score1"
# Higher scores indicate stronger collective expression of the candidate genes
summary(heart_seurat$candidate_score1)
```

# Step 4: Visualize Module Scores by Cell Type

## UMAP Visualization

The module score projected onto the UMAP embedding reveals which cell
populations show enriched expression of the candidate gene set.

```{r plot_umap, fig.cap = "UMAP colored by candidate gene module score. Warmer colors indicate higher collective expression."}
FeaturePlot(
  heart_seurat,
  features = "candidate_score1",
  cols = c("lightgrey", "blue", "red"),
  pt.size = 1.5
) +
  ggtitle("Candidate Gene Module Score") +
  theme(plot.title = element_text(hjust = 0.5))
```

## Violin Plot by Cell Type

A violin plot directly compares the distribution of module scores across
cell types.

```{r plot_violin, fig.cap = "Module score distribution by cell type. Cardiomyocytes show the highest enrichment."}
VlnPlot(
  heart_seurat,
  features = "candidate_score1",
  group.by = "cell_type",
  pt.size = 0
) +
  ggtitle("Candidate Gene Enrichment by Cell Type") +
  theme(axis.text.x = element_text(angle = 45, hjust = 1))
```

## Individual Gene Expression

Examining expression of individual candidate genes across cell types
identifies which candidates drive the module score.

```{r plot_individual_genes, fig.cap = "Dot plot of individual candidate genes by cell type. Size = percent expressing; color = mean expression."}
# Use only candidates present in the data
genes_to_plot <- intersect(cardiac_candidates, rownames(heart_seurat))

DotPlot(
  heart_seurat,
  features = genes_to_plot,
  group.by = "cell_type"
) +
  RotatedAxis() +
  ggtitle("Individual Candidate Gene Expression by Cell Type")
```

# Step 5: Quantify Cell-Type Specificity

Calculate summary statistics for candidate gene expression across cell types
to produce a quantitative ranking.

```{r quantify_specificity}
# Extract expression matrix for candidate genes
# Use layer= for Seurat v5 compatibility, fall back to slot= for v4
expr_matrix <- tryCatch(
  GetAssayData(heart_seurat, layer = "data"),
  error = function(e) GetAssayData(heart_seurat, slot = "data")
)
candidates_in_data <- intersect(cardiac_candidates, rownames(expr_matrix))

if (length(candidates_in_data) > 0) {
  # Calculate mean expression per cell type per gene
  cell_type_expr <- data.table(
    cell_type = heart_seurat$cell_type,
    as.data.table(t(as.matrix(expr_matrix[candidates_in_data, , drop = FALSE])))
  )

  # Aggregate by cell type
  mean_expr <- cell_type_expr[, lapply(.SD, mean), by = cell_type,
                               .SDcols = candidates_in_data]

  # Calculate percentage of cells expressing each gene (> 0) per cell type
  pct_expr <- cell_type_expr[, lapply(.SD, function(x) mean(x > 0) * 100),
                              by = cell_type, .SDcols = candidates_in_data]

  cat("Mean expression by cell type:\n")
  print(mean_expr)

  cat("\nPercent cells expressing (> 0) by cell type:\n")
  print(pct_expr)
}
```

# Step 6: Integrate Results into locusPackRat

The cell-type specificity results can be formatted and added to a
locusPackRat project as an additional evidence layer.

```{r add_to_project}
# Build a summary table of cell-type expression for each candidate
cell_type_names <- sort(unique(heart_seurat$cell_type))

specificity_summary <- data.table(
  gene_symbol = candidates_in_data,
  cardiomyocyte_expr = as.numeric(mean_expr[cell_type == "cardiomyocyte",
                                             ..candidates_in_data]),
  fibroblast_expr = as.numeric(mean_expr[cell_type == "fibroblast",
                                          ..candidates_in_data]),
  endothelial_expr = as.numeric(mean_expr[cell_type == "endothelial",
                                           ..candidates_in_data]),
  immune_expr = as.numeric(mean_expr[cell_type == "immune",
                                      ..candidates_in_data])
)

# Calculate a simple specificity index:
# ratio of cardiomyocyte expression to mean of all other cell types
specificity_summary[, cardiomyocyte_specificity :=
  cardiomyocyte_expr / rowMeans(.SD),
  .SDcols = c("fibroblast_expr", "endothelial_expr", "immune_expr")]

print(specificity_summary)

# Add to locusPackRat project
addRatTable(
  data = specificity_summary,
  table_name = "sc_cell_type_expression",
  abbreviation = "sc",
  link_type = "gene",
  link_by = "gene_symbol",
  project_dir = project_dir
)
```

## Project State After Integration

```{r show_project_state}
# Show all tables in the project
listPackRatTables(project_dir)
```

The single-cell evidence layer is now stored alongside any other annotations
in the project. It can be exported via `buildExcel()` or used for
downstream prioritization.

# Tips and Considerations

1. **Reference Dataset Selection**: Choose a scRNA-seq dataset from the tissue
   most relevant to your phenotype. Tabula Muris provides broad tissue coverage;
   tissue-specific atlases (e.g., the Human Cell Atlas) may offer finer resolution.

2. **Species Matching**: If your locusPackRat project uses mouse gene symbols,
   ensure the scRNA-seq reference also uses mouse symbols. For cross-species
   comparisons, use the orthology table generated by `initPackRat()`.

3. **Module Score Interpretation**: `AddModuleScore()` returns relative scores
   that control for average expression level. A high score means the gene set
   is more expressed than expected given the cells' overall transcriptional
   activity. Scores are most useful for comparing across cell types within the
   same dataset, not across datasets.

4. **Statistical Testing**: For formal statistical comparisons of module scores
   between cell types, consider using a Wilcoxon rank-sum test or permutation
   testing rather than relying solely on visual inspection.

5. **Scaling to Many Loci**: For studies with many candidate loci, you can
   create separate module scores for each locus and compare cell-type enrichment
   patterns across loci to identify shared or locus-specific cell type involvement.

6. **Single-Cell QTL Data**: As single-cell eQTL (sceQTL) datasets grow,
   consider querying these through `queryOpenTargetsQTL()` with
   `study_types = c("sceqtl", "scpqtl")` for variant-level cell-type-specific
   evidence that complements the expression-based approach shown here.

# Session Info

```{r session_info}
tryCatch(sessionInfo(), error = function(e) message("sessionInfo() unavailable"))
```

```{r cleanup, include = FALSE}
# Clean up the temporary project directory
reg.finalizer(environment(), function(e) {
  unlink(project_dir, recursive = TRUE, force = TRUE)
}, onexit = TRUE)
```
