---
title: "Single-Cell RNA-seq Integration for Cell-Type-Specific Candidate Evaluation"
author: "Brian Gural, Todd Kimball, Anh Luu, Christoph D. Rau"
date: "`r Sys.Date()`"
output:
    rmarkdown::html_vignette:
        toc: true
        toc_depth: 3
vignette: >
  %\VignetteIndexEntry{Single-Cell RNA-seq Integration for Cell-Type-Specific Candidate Evaluation}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  eval = requireNamespace("Seurat", quietly = TRUE) &&
         requireNamespace("ggplot2", quietly = TRUE) &&
         dir.exists("my_qtl_project/.locusPackRat")
)
```

# Overview

Bulk-tissue transcriptomic data, which form a common evidence layer in GWAS
candidate gene prioritization, can obscure effects that are specific to
particular cell types. A gene that appears broadly expressed in bulk tissue
may in fact be restricted to a small cell population, and conversely, a gene
with moderate bulk expression may be highly enriched in the disease-relevant
cell type.

This vignette demonstrates how to use single-cell RNA-seq (scRNA-seq) data
to evaluate candidate genes identified by `locusPackRat` for cell-type-specific
expression. Specifically, we show how to:

1. Load a reference scRNA-seq dataset (Tabula Muris)
2. Score candidate gene sets using Seurat's `AddModuleScore()`
3. Visualize module scores across cell types
4. Use the results to refine candidate gene prioritization

All code chunks are set to `eval=FALSE` since the scRNA-seq analysis requires
Seurat and reference data that may not be available in all environments.

# Prerequisites

```{r libraries}
# Core packages
library(locusPackRat)
library(data.table)

# scRNA-seq analysis
library(Seurat)     # install from CRAN: install.packages("Seurat")
library(ggplot2)
```

# Step 1: Extract Candidate Genes from locusPackRat

After running a locusPackRat workflow, extract the prioritized candidate genes
to evaluate for cell-type specificity.

```{r extract_candidates}
# Load the master gene sheet from a completed locusPackRat project
# This assumes you have already run initPackRat(), addRatTable(), etc.
master_genes <- fread("my_qtl_project/.locusPackRat/input/genes.csv")

# Define candidate gene sets based on your prioritization criteria
# These are example genes from the CC cardiac hypertrophy study
cardiac_candidates <- c("Cisd2", "Pdlim5", "Manba", "Fhod3", "Myh7")

# You can also define broader gene sets for comparison
all_locus_genes <- master_genes$gene_symbol
```

# Step 2: Load a Reference scRNA-seq Dataset

We use the Tabula Muris dataset as a reference atlas of mouse cell types.
The Tabula Muris Senis (aged) dataset includes heart tissue with annotated
cardiomyocytes, fibroblasts, endothelial cells, and immune cells.

## Option A: Load from Tabula Muris (10X Genomics data)

```{r load_tabula_muris}
# Download Tabula Muris heart data (10X Genomics)
# Data available from: https://tabula-muris.ds.czbiohub.org/
# Or via the TabulaMurisData Bioconductor package

# Using pre-downloaded .h5 or .rds file:
heart_data <- Read10X(data.dir = "path/to/tabula_muris/Heart-10X_P7_4/")

# Create Seurat object
heart_seurat <- CreateSeuratObject(
  counts = heart_data,
  project = "TabulaMuris_Heart",
  min.cells = 3,
  min.features = 200
)

# Standard preprocessing
heart_seurat <- NormalizeData(heart_seurat)
heart_seurat <- FindVariableFeatures(heart_seurat, nfeatures = 2000)
heart_seurat <- ScaleData(heart_seurat)
heart_seurat <- RunPCA(heart_seurat)
heart_seurat <- FindNeighbors(heart_seurat, dims = 1:20)
heart_seurat <- FindClusters(heart_seurat, resolution = 0.5)
heart_seurat <- RunUMAP(heart_seurat, dims = 1:20)
```

## Option B: Load Pre-Processed Seurat Object

If you have a pre-processed Seurat object with cell type annotations:

```{r load_preprocessed}
# Load a pre-processed Seurat object with cell type annotations
heart_seurat <- readRDS("path/to/heart_annotated_seurat.rds")

# Check available cell type annotations
table(heart_seurat$cell_type)
```

# Step 3: Score Candidate Genes with AddModuleScore

Seurat's `AddModuleScore()` calculates per-cell enrichment scores for a set
of genes, controlling for expression level and gene set size. This allows us
to ask: "Are our candidate genes collectively enriched in specific cell types?"

```{r module_scores}
# Define gene lists for module scoring
# Convert mouse gene symbols to match the Seurat object's feature names
candidate_list <- list(
  cardiac_candidates = cardiac_candidates
)

# Calculate module scores
heart_seurat <- AddModuleScore(
  object = heart_seurat,
  features = candidate_list,
  name = "candidate_score"
)

# The score is added as a metadata column: "candidate_score1"
# Higher scores indicate stronger collective expression of the candidate genes
summary(heart_seurat$candidate_score1)
```

# Step 4: Visualize Module Scores by Cell Type

## UMAP Visualization

```{r plot_umap}
# Plot module score on UMAP
FeaturePlot(
  heart_seurat,
  features = "candidate_score1",
  cols = c("lightgrey", "blue", "red"),
  pt.size = 0.5
) +
  ggtitle("Candidate Gene Module Score") +
  theme(plot.title = element_text(hjust = 0.5))
```

## Violin Plot by Cell Type

```{r plot_violin}
# Violin plot of module score by cell type
VlnPlot(
  heart_seurat,
  features = "candidate_score1",
  group.by = "cell_type",
  pt.size = 0
) +
  ggtitle("Candidate Gene Enrichment by Cell Type") +
  theme(axis.text.x = element_text(angle = 45, hjust = 1))
```

## Individual Gene Expression

Examine expression of individual candidate genes across cell types to
identify which candidates drive the module score.

```{r plot_individual_genes}
# Dot plot of individual candidate gene expression by cell type
DotPlot(
  heart_seurat,
  features = cardiac_candidates,
  group.by = "cell_type"
) +
  RotatedAxis() +
  ggtitle("Individual Candidate Gene Expression by Cell Type")
```

# Step 5: Quantify Cell-Type Specificity

Calculate summary statistics for candidate gene expression across cell types
to produce a quantitative ranking.

```{r quantify_specificity}
# Extract expression matrix for candidate genes
expr_matrix <- GetAssayData(heart_seurat, slot = "data")
candidates_in_data <- intersect(cardiac_candidates, rownames(expr_matrix))

if (length(candidates_in_data) > 0) {
  # Calculate mean expression per cell type per gene
  cell_type_expr <- data.table(
    cell_type = heart_seurat$cell_type,
    as.data.table(t(as.matrix(expr_matrix[candidates_in_data, , drop = FALSE])))
  )

  # Aggregate by cell type
  mean_expr <- cell_type_expr[, lapply(.SD, mean), by = cell_type,
                               .SDcols = candidates_in_data]

  # Calculate percentage of cells expressing each gene (> 0) per cell type
  pct_expr <- cell_type_expr[, lapply(.SD, function(x) mean(x > 0) * 100),
                              by = cell_type, .SDcols = candidates_in_data]

  cat("Mean expression by cell type:\n")
  print(mean_expr)

  cat("\nPercent cells expressing (> 0) by cell type:\n")
  print(pct_expr)
}
```

# Step 6: Integrate Results into locusPackRat

The cell-type specificity results can be formatted and added to a
locusPackRat project as an additional evidence layer.

```{r add_to_project}
# Create a summary table of cell-type specificity for each candidate
specificity_summary <- data.table(
  gene_symbol = candidates_in_data,
  cardiomyocyte_expr = as.numeric(mean_expr[cell_type == "cardiomyocyte",
                                             ..candidates_in_data]),
  fibroblast_expr = as.numeric(mean_expr[cell_type == "fibroblast",
                                          ..candidates_in_data]),
  endothelial_expr = as.numeric(mean_expr[cell_type == "endothelial cell",
                                           ..candidates_in_data]),
  immune_expr = as.numeric(mean_expr[cell_type == "immune cell",
                                      ..candidates_in_data])
)

# Calculate a simple specificity index
# Ratio of target cell type expression to mean of all other cell types
specificity_summary[, cardiomyocyte_specificity :=
  cardiomyocyte_expr / rowMeans(.SD),
  .SDcols = c("fibroblast_expr", "endothelial_expr", "immune_expr")]

# Add to locusPackRat project
addRatTable(
  data = specificity_summary,
  table_name = "sc_cell_type_expression",
  abbreviation = "sc",
  link_type = "gene",
  link_by = "gene_symbol",
  project_dir = "my_qtl_project"
)
```

# Tips and Considerations

1. **Reference Dataset Selection**: Choose a scRNA-seq dataset from the tissue
   most relevant to your phenotype. Tabula Muris provides broad tissue coverage;
   tissue-specific atlases (e.g., the Human Cell Atlas) may offer finer resolution.

2. **Species Matching**: If your locusPackRat project uses mouse gene symbols,
   ensure the scRNA-seq reference also uses mouse symbols. For cross-species
   comparisons, use the orthology table generated by `initPackRat()`.

3. **Module Score Interpretation**: `AddModuleScore()` returns relative scores
   that control for average expression level. A high score means the gene set
   is more expressed than expected given the cells' overall transcriptional
   activity. Scores are most useful for comparing across cell types within the
   same dataset, not across datasets.

4. **Statistical Testing**: For formal statistical comparisons of module scores
   between cell types, consider using a Wilcoxon rank-sum test or permutation
   testing rather than relying solely on visual inspection.

5. **Scaling to Many Loci**: For studies with many candidate loci, you can
   create separate module scores for each locus and compare cell-type enrichment
   patterns across loci to identify shared or locus-specific cell type involvement.

6. **Single-Cell QTL Data**: As single-cell eQTL (sceQTL) datasets grow,
   consider querying these through `queryOpenTargetsQTL()` with
   `study_types = c("sceqtl", "scpqtl")` for variant-level cell-type-specific
   evidence that complements the expression-based approach shown here.

# Session Info

```{r session_info}
sessionInfo()
```
