---
title: "Querying GeneNetwork2 for Model Organism QTL Data"
author: "Brian Gural, Todd Kimball, Anh Luu, Christoph D. Rau"
date: "`r Sys.Date()`"
output:
    rmarkdown::html_vignette:
        toc: true
        toc_depth: 3
vignette: >
  %\VignetteIndexEntry{Querying GeneNetwork2 for Model Organism QTL Data}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  eval = TRUE
)
```

# Overview

This vignette demonstrates how to query the [GeneNetwork2](https://genenetwork.org)
REST API to retrieve eQTL and phenotype data from model organism genetic reference
populations (GRPs), and how to integrate those results into a `locusPackRat` project
as additional evidence layers.

GeneNetwork2 hosts QTL mapping results, expression data, and phenotype records for
several GRPs, including:

- **BxD recombinant inbred lines** (mouse): extensive eQTL, pQTL, and phenotype data
- **Diversity Outbred (DO)** (mouse): high-resolution QTL mapping panels
- **Hybrid Mouse Diversity Panel (HMDP)**: cardiovascular and metabolic trait data
- **Other organisms**: rat (HxB/BxH), Drosophila, and Arabidopsis panels

Unlike Open Targets, which focuses on human GWAS data, GeneNetwork2 provides
population-specific molecular QTL data that is directly relevant to model organism
studies. This makes it a valuable complement to the human-centric annotations
retrieved by `queryOpenTargets()` and `queryOpenTargetsQTL()`.

Live API calls are attempted first; if the GeneNetwork2 API is unavailable, cached
responses bundled with the package are used as a fallback.

# Prerequisites

```{r libraries}
library(locusPackRat)
library(data.table)
library(httr)
library(jsonlite)
```

# GeneNetwork2 API Basics

The GeneNetwork2 REST API is available at `https://genenetwork.org/api/v_pre1/`.
Key endpoints include:

- **`/species`**: List available species
- **`/groups/{species}`**: List available populations (e.g., BXD, DO)
- **`/datasets/{group}`**: List available datasets for a population
- **`/traits/{dataset}`**: List traits within a dataset
- **`/trait/{dataset}/{probe_name}`**: Get detailed trait information

```{r api_setup}
# Base URL for GeneNetwork2 API
gn2_base <- "https://genenetwork.org/api/v_pre1"

# Helper: fetch from API with cached fallback
gn_fetch_or_cache <- function(url, cache_file, timeout_sec = 30) {
  result <- tryCatch({
    resp <- httr::GET(url, httr::timeout(timeout_sec))
    httr::stop_for_status(resp)
    jsonlite::fromJSON(httr::content(resp, as = "text", encoding = "UTF-8"))
  }, error = function(e) NULL)

  if (!is.null(result) && length(result) > 0) {
    message("GeneNetwork API: live data retrieved")
    return(result)
  }

  cache_path <- system.file("extdata", "gn_cache", cache_file,
                            package = "locusPackRat")
  if (nzchar(cache_path) && file.exists(cache_path)) {
    message("GeneNetwork API: using cached data (API unavailable)")
    return(jsonlite::fromJSON(cache_path))
  }
  stop("GeneNetwork API unavailable and no cached data found for: ", cache_file)
}
```

# Example 1: BxD eQTL Data

## List Available BxD Expression Datasets

The BxD panel has dozens of expression datasets spanning multiple tissues
(brain, liver, heart, kidney, etc.) and platforms (microarray, RNA-seq).

```{r list_bxd_datasets}
# Fetch all datasets for the BXD group
bxd_datasets <- gn_fetch_or_cache(
  paste0(gn2_base, "/datasets/BXD"),
  "bxd_datasets.json"
)

# The FullName column contains tissue and platform info
# Filter for expression datasets related to heart/cardiac tissue
heart_datasets <- bxd_datasets[grepl("Heart|Cardiac|heart|cardiac",
                                     bxd_datasets$FullName), ]

cat("Total BXD datasets:", nrow(bxd_datasets), "\n")
cat("Heart-related datasets:", nrow(heart_datasets), "\n\n")

# Preview heart datasets
display_cols <- intersect(names(heart_datasets),
                          c("Id", "FullName", "Short_Abbreviation"))
head(heart_datasets[, display_cols, drop = FALSE])
```

## Query Traits for a Specific Gene

To check whether a candidate gene has eQTL evidence in the BxD panel,
query traits by gene symbol within a specific dataset. The traits endpoint
returns all probes/traits; we filter to our gene of interest.

```{r query_gene_traits}
# Query Cisd2 expression traits in a BXD heart dataset
dataset_name <- "HC_M2_0606_P"
gene_of_interest <- "Cisd2"

# Fetch traits (pre-filtered to Cisd2 in cached version)
cisd2_traits <- gn_fetch_or_cache(
  paste0(gn2_base, "/traits/", dataset_name),
  "cisd2_traits.json"
)

# The live endpoint returns ALL traits; filter to our gene
# The cached version is already pre-filtered
if (nrow(cisd2_traits) > 20) {
  # Live data -- filter by Symbol column
  cisd2_traits <- cisd2_traits[grepl(gene_of_interest, cisd2_traits$Symbol,
                                     ignore.case = TRUE), ]
}

cat("Found", nrow(cisd2_traits), "probes/traits for", gene_of_interest, "\n\n")
print(cisd2_traits[, c("Name", "Symbol", "Description", "Chr", "Mb")])
```

Each row represents a microarray probe targeting Cisd2. The `LRS` column gives
the likelihood ratio statistic for the best eQTL, and `Peak Chr` / `Peak Mb`
indicate the genomic position of the peak association.

## Retrieve Trait Detail

Get detailed information for a specific probe, including the additive effect
and peak association statistics.

```{r get_trait_values}
# Get detail for the first (strongest) Cisd2 probe
probe_name <- cisd2_traits$Name[1]

trait_detail <- gn_fetch_or_cache(
  paste0(gn2_base, "/trait/", dataset_name, "/", probe_name),
  "cisd2_trait_detail.json"
)

cat("Probe:", trait_detail$name, "\n")
cat("Gene:", trait_detail$symbol, "\n")
cat("Description:", trait_detail$description, "\n")
cat("Chromosome:", trait_detail$chr, "at", trait_detail$mb, "Mb\n")
cat("Best LRS:", round(trait_detail$lrs, 2),
    "(LOD ~", round(trait_detail$lrs / 4.61, 2), ")\n")
cat("Peak locus:", trait_detail$locus, "\n")
cat("Additive effect:", round(trait_detail$additive, 4), "\n")
```

## Run QTL Mapping via GEMMA

GeneNetwork2 can perform on-the-fly QTL mapping using GEMMA (Genome-wide
Efficient Mixed Model Association). This returns LOD scores across the genome.

```{r run_gemma}
# Request GEMMA mapping for the Cisd2 probe
# Note: this endpoint may take 30-60 seconds to return results
gemma_url <- paste0(gn2_base, "/mapping?trait_id=", probe_name,
                    "&db=", dataset_name,
                    "&method=gemma")

gemma_raw <- gn_fetch_or_cache(gemma_url, "cisd2_gemma.json", timeout_sec = 120)

# The API returns a nested list; extract the data frame
if (is.data.frame(gemma_raw)) {
  gemma_results <- gemma_raw
} else if (is.list(gemma_raw) && length(gemma_raw) > 0) {
  gemma_results <- gemma_raw[[1]]
} else {
  gemma_results <- data.frame()
}

if (nrow(gemma_results) > 0) {
  eqtl_scan <- data.table(
    chr = gemma_results$chr,
    pos = gemma_results$Mb * 1e6,  # Convert Mb to bp
    pos_mb = gemma_results$Mb,
    lod = gemma_results$lod_score,
    p_value = gemma_results$p_value,
    marker = gemma_results$name,
    gene_symbol = gene_of_interest,
    source = "GeneNetwork2_BxD"
  )

  # Summary of the scan
  cat("Total markers scanned:", nrow(eqtl_scan), "\n")
  cat("Significant positions (LOD > 3):", sum(eqtl_scan$lod > 3), "\n\n")

  # Show the top peaks
  top_peaks <- eqtl_scan[order(-lod)][1:5]
  print(top_peaks[, .(chr, pos_mb, lod, p_value, marker)])
}
```

The strongest eQTL for Cisd2 maps to chromosome 3, near the gene's own
position -- consistent with a *cis*-eQTL, which supports its candidacy in a
QTL study targeting this region.

# Example 2: BXD Published Phenotypes

GeneNetwork2 also hosts curated phenotype data for BXD strains. Published
phenotypes can be accessed through the `BXDPublish` dataset, providing
higher-level trait-QTL associations.

```{r list_bxd_phenotypes}
# Fetch BXD published phenotype traits
bxd_pheno <- tryCatch({
  gn_fetch_or_cache(
    paste0(gn2_base, "/traits/BXDPublish"),
    cache_file = NULL  # no cache for this endpoint
  )
}, error = function(e) {
  message("BXD published phenotypes unavailable: ", e$message)
  NULL
})

if (!is.null(bxd_pheno) && nrow(bxd_pheno) > 0) {
  cat("Total BXD published phenotype traits:", nrow(bxd_pheno), "\n\n")

  # Show a few example traits
  display_cols <- intersect(names(bxd_pheno), c("Id", "Name", "Description", "Symbol"))
  print(head(bxd_pheno[, display_cols, drop = FALSE], 10))
} else {
  cat("BXD published phenotype data not available via this API endpoint.\n")
  cat("Visit https://genenetwork.org to browse phenotypes interactively.\n")
}
```

# Integrating GeneNetwork2 Results into locusPackRat

## Create a locusPackRat Project

```{r create_project}
# Create a temporary project with candidate genes
project_dir <- file.path(tempdir(), "gn_demo_project")

gene_input <- data.frame(
  gene_symbol = c("Cisd2", "Pdlim5", "Manba", "Fhod3", "Myh7")
)

initPackRat(
  data = gene_input,
  mode = "gene",
  species = "mouse",
  genome = "mm39",
  project_dir = project_dir,
  force = TRUE
)
```

## Gene-Level Integration

Use `link_type = "gene"` when you have per-gene summary statistics
(e.g., peak LOD score, cis/trans classification). Here we build a summary
from the GEMMA results computed above.

```{r integrate_gene_level}
# Build a gene-level summary from the GEMMA scan
# In a real analysis, you would loop over multiple candidate genes
if (exists("eqtl_scan") && nrow(eqtl_scan) > 0) {
  peak_row <- eqtl_scan[which.max(lod)]
  gene_chr <- unique(eqtl_scan[chr == peak_row$chr, chr])

  eqtl_summary <- data.table(
    gene_symbol = gene_of_interest,
    bxd_max_lod = round(peak_row$lod, 2),
    bxd_peak_chr = as.character(peak_row$chr),
    bxd_peak_mb = round(peak_row$pos_mb, 2),
    bxd_cis_eqtl = (as.character(peak_row$chr) == as.character(cisd2_traits$Chr[1])),
    gn2_dataset = dataset_name
  )

  print(eqtl_summary)

  # Add to locusPackRat project
  addRatTable(
    data = eqtl_summary,
    table_name = "bxd_eqtl",
    abbreviation = "bxd",
    link_type = "gene",
    link_by = "gene_symbol",
    project_dir = project_dir
  )
}
```

## Point-Level Integration

Use `link_type = "region"` when you have position-level QTL scan data
that should be mapped to overlapping genes/regions.

```{r integrate_point_level}
# Use the full scan data from GEMMA -- keep significant hits only
if (exists("eqtl_scan") && nrow(eqtl_scan) > 0) {
  significant_hits <- eqtl_scan[lod > 3]
  cat("Integrating", nrow(significant_hits), "significant markers\n")

  scan_for_project <- significant_hits[, .(
    chr = as.character(chr),
    start = as.integer(pos),
    end = as.integer(pos),
    lod = round(lod, 3),
    gene_symbol = gene_symbol,
    source = source
  )]

  addRatTable(
    data = scan_for_project,
    table_name = "bxd_eqtl_scan",
    abbreviation = "bxds",
    link_type = "region",
    link_by = "chr,start,end",
    project_dir = project_dir
  )
}
```

## Project State After Integration

```{r show_project_state}
listPackRatTables(project_dir)
```

# Alternative: GNapi R Package

The [GNapi](https://github.com/kbroman/GNapi) R package by Karl Broman
provides a more structured R client for the GeneNetwork2 API. If you
prefer a packaged interface:

```{r gnapi_example, eval = FALSE}
# Install GNapi (https://github.com/kbroman/GNapi)
# remotes::install_github("kbroman/GNapi")
library(GNapi)

# List available species and groups
species <- list_species()
groups  <- list_groups("mouse")

# List BxD datasets
bxd_datasets <- list_datasets("BXD")

# Run GEMMA QTL scan for a specific trait
gemma_scan <- run_gemma("HC_M2_0606_P", "1428441_at")

# Find correlated traits
corr <- run_correlation("HC_M2_0606_P", "BXD", "1428441_at")
```

# Tips and Considerations

1. **API Rate Limits**: GeneNetwork2 may impose rate limits. Add `Sys.sleep(1)`
   between requests when querying many genes.

2. **Genome Build**: GeneNetwork2 positions are typically in GRCm38/mm10 for
   mouse data. If your locusPackRat project uses mm39, convert coordinates
   using `rtracklayer::liftOver()` before integration (see the liftOver
   section in the main workflow vignette).

3. **Dataset Selection**: Choose tissue-appropriate datasets. For a cardiac
   study, prioritize heart expression datasets over brain or liver.

4. **Cis vs Trans**: A cis-eQTL (peak near the gene itself) is generally
   stronger evidence for candidacy than a trans-eQTL.

5. **Cross-Validation**: Compare GeneNetwork2 eQTL results with your own
   study-specific eQTL data and Open Targets QTL data for convergent evidence.

# Session Info

```{r session_info}
sessionInfo()
```

```{r cleanup, include = FALSE}
reg.finalizer(environment(), function(e) {
  unlink(project_dir, recursive = TRUE, force = TRUE)
}, onexit = TRUE)
```
